---
title: "1 Data wrangling and filtering"
author: "Sandra Erdmann"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: inline
execute:
  freeze: auto
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

# Abbreviations

Reef names:



-   BRR = Bramble Reef

-   EP = East Pelorus

-   GB = Geoffrey Bay

-   HaR = Havannah Reef

-   HB = Horseshoe Bay

-   HFB = Huntingfield Bay

-   JB = John Brewer Reef

-   KR = Keeper Reef

-   LPB = Little Pioneer Bay

-   MB = Maud Bay

-   MR = Middle Reef

-   PB = Picnic Bay

-   SO = South Orpheus

-   WB = Wilson Bay

-   WP = West Pelorus

Data sets:

ak = Acropora kenti SNP dataset ak.filtered = used ak, which has been filtered for all downstream analysis ak.filtered.nr = used ak.filtered and removed replicates (nr = no replicates) ak.filtered.nr.nm = = used ak.filtered.nr and removed migrants (nm = no migrants) ak.pop = used ak.filtered.nr to run population structure analysis ak.pop.nm = used ak.pop and removed migrants ak.pop.mag = population structure dataset for Magnetic Island only ak.rel = used ak.filtered.nr.nm to run relatedness analysis ak.gen = used ak.filtered.nr.nm to calculate genetic diversity estimates (Fst, Fis, He, Ho) ak.gen.mag = used ak.gen to calculate genetic diversity metrics for Magnetic Island only ak.gen.adj = used ak.gen to calculate genetic diversity metrics for adjacent reefs only


```{r, eval=FALSE}
# Installation of dartR

# See https://green-striped-gecko.github.io/kioloa/install.html

# Install the necessary bioconductor packages
install.packages("devtools")
install.packages("BiocManager")
BiocManager::install("SNPRelate")

# Install dartRverse (dartRverse) & core (dartR.base, dartR.data)
install.packages("dartRverse")
library("dartRverse")
```

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("snpStats")
```

```{r, eval=FALSE}
# Install dartR packages
install.packages("dartR.sim")
install.packages("dartR.popgen")
install.packages("dartR.spatial")
install.packages("dartR.captive")

# Update dartR.base
dartRverse_install(package = "dartR.base", rep = "github", branch = "dev")
# Check which packages and versions are installed
dartRverse_install()
```


```{r}
## Load library
library(dartRverse)
```

# Load and inspect raw dataset

The SNP input dart file is named Report_DAc23-8597_5_moreOrders_SNP_2.csv. The metadata file is called "ind_metrics_SNP2.csv" and contains information for all individuals, such as, ID, name of populations, locations and coordinates of sample locations.

```{r}
# Merge the DartR SNP dataset with the metadata file
if (file.exists("cache/ak.rdata")) {
  load("cache/ak.rdata")
} else {

  ak <- gl.read.dart(filename="raw_data/Report_DAc23-8597_5_moreOrders_SNP_2.csv", ind.metafile = "ind_metrics_SNP2.csv")
 # Save merged SNP and metadata file
  save(ak, file="cache/ak.rdata")    
}
```

```{r}
# Get names of individuals, location and population
nInd(ak)
nLoc(ak)
nPop(ak)
```

```{r}
# Create a table showing the number of individuals per population
table(pop(ak))
```

```{r}
# Create a barplot from that table
barplot(table(pop(ak)), las=2)
```

```{r}
# Run a smearplot to assess quality of dataset
gl.smearplot(ak)
```

# Assess callrate

```{r}
# Get names of individual samples
ak$ind.names
```

Inspect the individual call rates and remove duplicates with the lower call rate

```{r}
gl.report.callrate(ak, method = "ind", ind.to.list = 528)
```

# Remove duplicates

Remove the following duplicates: At_162_J20.1', 'At_163_J20.1', 'B1', 'A2', 'A3', 'A7', 'B8', 'A9', 'A10', 'B11', 'A15', 'B17', 'B18', 'A19', 'A21', 'WB_03/04/2022_36'

```{r}

if (file.exists("cache/ak1.rdata")){
  load("cache/ak1.rdata")
} else {
  # Drop duplicates from current dataset
  ak1 <- gl.drop.ind(ak, ind.list=c('At_162_J20.1', 'At_163_J20.1', 'B1', 'A2', 'A3', 'A7', 'B8', 'A9', 'A10', 'B11', 'A15', 'B17', 'B18', 'A19', 'A21', 'WB_03/04/2022_36'))
  # Filter for monomorphs after removing duplicates}
  ak1 <- gl.filter.monomorphs(ak1)
  # Recalculate locus metrics
  ak1 <- gl.recalc.metrics(ak1)
  save(ak1,file = "cache/ak1.rdata")
}
```

# Filtering

Filter dataset for

1.  reproducibility 0.99
2.  read depth \>10 and \<100
3.  maf (0.005)
4.  call rate for loci (0.8)
5.  call rate for individuals (0.75)
6.  monomorphs
7.  secondaries
8.  missing loci.

```{r}
nInd(ak1)
nLoc(ak1)
```

### Reproducibility 0.99

A metrics that investigates if calling a SNP twice results in the same outcome.

```{r}
# Inspect the current reproducibility to remove unreliable loci
gl.report.reproducibility(ak1)
```

```{r}
# Filter loci for reproducibility due to high missing rate
ak2 <- gl.filter.reproducibility(ak1)
```

```{r}
nInd(ak2)
nLoc(ak2)
```

### Read depth \>10 \<100

Loci with high (= unreliable) and low read depth (= non-confident) should be removed.

Inspect the current read depth

```{r}
gl.report.rdepth(ak2)
```

Filter for a read depth greater than 10 and lower than 100.

```{r}
ak3 <- gl.filter.rdepth(ak2, lower = 10, upper = 100)
```

```{r}
nInd(ak3)
nLoc(ak3)
```

### Minor allele frequency (maf) 0.005

With approximately 500 individual samples this filter will remove variant loci supported by a minor allele count of just 1 or 2.

```{r}
# Inspect the current minor allele frequency
gl.report.maf(ak3)
```

Remove sites with a minor allele frequency of less than 0.005

```{r}
ak4 <- gl.filter.maf(ak3,  threshold = 0.005)
```

```{r}
nInd(ak4)
nLoc(ak4)
```

### Call rate for loci 0.8

Retaining only loci that are consistently called across samples (at least 80%)

```{r}
# Inspect the current call rate for loci
gl.report.callrate(ak4)
```

```{r}
# Filter fora call rate of 0.8
ak5 <- gl.filter.callrate(ak4, method = "loc", threshold = 0.8)
```

```{r}
nInd(ak5)
nLoc(ak5)
```

### Call rate for individual 0.75

Most individuals have high callrates (\>80% across all loci) but there is a cluster of very poorly sequenced individuals with low call rates. We remove these individuals from the dataset as they have sequenced poorly.

```{r}
# Inspect the current callrate for individuals
gl.report.callrate(ak5, method="ind")
```

```{r}
# Filter for a call rate for individuals of 0.75
ak6 <-gl.filter.callrate(ak5, method ="ind", threshold=0.75)
```

```{r}
# Recalculate locus metrics
ak6 <- gl.recalc.metrics(ak6)
```

```{r}
nInd(ak6)
nLoc(ak6)
```

### Monomorphs

This step deletes monomorphic loci from a genlight {adegenet} object. A DArT dataset will not have monomorphic loci, but they can arise, along with loci when populations or individuals are deleted.

```{r}
# Inspect the current number of monomorphs
gl.report.monomorphs(ak6)
```

```{r}
# Filter for monomorphs
ak7 <- gl.filter.monomorphs(ak6 ,verbose = 5)
```

```{r}
nInd(ak7)
nLoc(ak7)
```

### Secondaries

Secondaries are fragments with more than one SNP. These multiple SNP loci within a fragment are likely to be linked, and so you may wish to remove secondaries. The filter removes loci that may not be inherited independently. Filtering secondaries at the end of the filtering process ensures that highly quality loci remain in the dataset.

```{r}
# Inspect the current number of secondaries
gl.report.secondaries(ak7)
```

```{r}
# Filter for secondaries
ak8 <- gl.filter.secondaries(ak7, verbose = 5)
```

```{r}
nInd(ak8)
nLoc(ak8)
```

### Remove missing loci

A DArT dataset will not have individuals for which all calls are scored as missing (NA) across all loci, but such individuals may sneak in to the dataset when loci are deleted. Retaining individual or loci with all NAs can cause issues for several functions.

```{r}
# Remove missing loci
ak9 <-gl.filter.allna(ak8)
```

```{r}
nInd(ak9)
nLoc(ak9)
```

```{r}
# Run a smearplot to assess quality of dataset
gl.smearplot(ak9)
```

```{r}
# Recalculate locus metrics
ak9 <- gl.recalc.metrics(ak9)
```

# Inspect filtered dataset

```{r}
# A compliance check ensures the dataset has no monomorphic loci, no missing data, properly assigned populations, individual names are unique, coordinates are correct, and recalculates locus metrics.
gl.compliance.check(ak9)
```

```{r}
nInd(ak9)
nLoc(ak9)
nPop(ak9)
```

```{r}
# Create a table showing the number of individuals per population
table(pop(ak9))
```

```{r}
# Create a barplot from that table
barplot(table(pop(ak9)), las=2)
```

```{r}
# rename the dataset after the filtering process is completed
ak.filtered <- ak9
```

```{r}
# Get names of individuals
ak.filtered$ind.names
```

```{r}
# Rename sample "20" to WB_36 for consistency
ak.filtered$ind.names[ak.filtered$ind.names == "20"] <- "WB_36"
```

```{r}
# Confirm name change
ak.filtered$ind.names
```

```{r}
# Save filtered and clean dataset
save(ak.filtered, file="cache/ak.filtered.rdata")
```

## Create a vcf file using identical filters to dartR

This conversion is done using bash scripts in the `dart2vcf` directory. These scripts require a list of loci and individuals remaining after our dart filtering process so we export those here.

```{r}
write.table(ak.filtered$loc.names,file = "dart2vcf/ak.filtered.loci",row.names = FALSE,quote = FALSE)
write.table(ak.filtered$ind.names,file = "dart2vcf/ak.filtered.ind",row.names = FALSE,quote = FALSE,col.names = FALSE)
```

The dart2vcf script is then run against the Acropora kenti genome, aligning dart tags to the genome and using this to obtain genomic coordinates for all tags.

``` bash
cd dart2vcf
./dart2vcf.py -g aten_final_0.11.fasta ../raw_data/Report_DAc23-8597_5_moreOrders_SNP_2.csv > 8597.vcf 2>8597.err
```

This process results in a loss of around 10% of loci (50323 of 55247), mostly because they cannot be mapped to the genome. Thus the vcf file will have slighly fewer loci than the genlight object used by dartR.

After performing the conversion we extract a subset corresponding to the same SNPs and individuals as `ak.filtered`

```bash
# Sanitize loci because dartR changes them from the original
#
cat ak.filtered.loci| sed 's/-/_/' | sed 's:/:>:' | sed 's/-/:/' | sed 's/_/|F|0--/'  > tmp.loci 

# Retains 4919 of the original 5276 loci
bcftools filter -i "ID=@tmp.loci" 8597.vcf.gz -Oz > 8597.1.vcf.gz

# Retains all 465 samples
bcftools view -S ak.filtered.ind 8597.1.vcf.gz -Oz > ak.filtered.vcf.gz
```
