---
title: "3 Population structure"
author: "Sandra Erdmann and Ira Cooke"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: inline
execute:
  freeze: auto
---

```{r, include=FALSE}
## Load Libraries
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = FALSE)

library(tidyverse)
library(dartRverse)
library(dartR.base)
library(ggpubr)
library(sf)
library(ozmaps)
source("constants.R")
library(ggrepel)

options(readr.show_col_types = FALSE)
```

For population structure we begin by loading the filtered data with redundant individuals removed.

```{r, echo=TRUE}
load("cache/ak.filtered.nr.rdata")
```

# Analysis of broadscale structure using the full dataset

We begin by examining the full filtered dataset, including data from both Magnetic Island and adjacent reefs. The data is first converted to structure format to run externally with structure program.

```{r, eval=FALSE, echo=TRUE}
gl2structure(ak.filtered.nr, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.filtered.nr.str", outpath = "structure", verbose = NULL)
```

Structure was then run for K = 1, 2, 3 and 4. Each run used 20000 MCMC iterations with 10000 additional as burn-in. Full parameter files can be found in the `structure` folder

Determining the appropriate value of K is tricky. The structure manual recommends a procedure based on finding the maximum likelihood value of K but also notes that care must be taken with this approach and especially cautions against selecting high values of K without a realistic biological explanation.

We employed the more objective method developed by [Evanno 2005](doi.org/10.1111/j.1365-294X.2005.02553.x) which involves taking the second derivative of the log Likelihood to calculate a statistic which they refer to as $\Delta K$. This method shows a very clear result, with the maximum value of $\Delta K$ achieved for $K=2$.

```{r}
llpaths <- list.files("structure","k_vs_lik.txt",recursive = TRUE,full.names = TRUE)
reps <- llpaths %>% str_extract("([0-9]+)")

read_loglik <- function(path,n){
  read_table(path,col_names = c("k","logLik"))  %>%  
    mutate(Ldash = logLik - lag(logLik,1)) %>% 
    mutate(deltaL = lead(Ldash) - Ldash) %>% 
    add_column(rep=n)
}


lldata <- map2(llpaths,reps,read_loglik) %>% 
  list_rbind()

lldata %>% 
  group_by(k) %>% 
  summarise(deltaK = mean(abs(deltaL))/var(logLik)) %>% 
  na.omit() %>% 
  ggplot(aes(x=k,y=deltaK)) + 
    geom_point() + geom_line() + xlab("Number of clusters, K") + ylab(expression(paste(Delta,"K")))
```

## Admixture plot for K=2

To construct an admixture plot for K=2 we read assignment probabilities from the structure outputs.

```{r, echo=TRUE}
# Load csv with data from run # 12 in the structure analysis
run12_samples <- read_table("structure/ak.filtered.nr.samples.txt",col_names = c("ID"))

run12 <-  read_table("structure/ak.structure.nr.k2.out.anc.txt",col_names = c("X1", "Label","(%Miss)","C1","C2"),skip = 2) %>% 
  cbind(run12_samples) %>% 
  select(ID,C1,C2) %>% # Select only useful columns
  pivot_longer(-ID,names_to = "cluster",values_to = "p") # Put assignment probabilities into a single column
```

We then join structure data with sample metadata

```{r, echo=TRUE}
metadata <- read_csv("ind_metrics_SNP2.csv") %>% 
  mutate(ID = case_when(
    id=="20" ~ "WB_36",
    .default = id
  )) %>% 
  select(-id)
save(metadata,file = "cache/metadata.rdata")


run12_meta <- run12 %>% 
  left_join(metadata,by="ID")
```

In order to make sense of population structure it is useful to visualise sampling locations on a map.

```{r}
sites_data <- run12_meta %>% 
  group_by(pop,lat,lon) %>% 
  summarise(n_samples = n()/2) %>% 
  mutate(name = location_names[pop])
```

```{r, include=FALSE}
gbr_coast <- st_read("maps/GBR_NESP-TWQ-5/BaseLayers/",layer = "GBR_coastlineONLY")
gbr_tc <- st_read("maps/GBR_NESP-TWQ-5/BaseLayers/",layer = "GBR_towncities")
gbr_features <- st_read("maps/GBR_NESP-TWQ-5/BaseLayers/",layer = "TS_AIMS_NESP_Torres_Strait_Features_V1b_with_GBR_Features")

sf_use_s2(FALSE) # the input geometries are bad, and wrong spherically
x1 <- st_make_valid(gbr_features)

xmn=146.0
ymn=-19.3
xmx=146.9
ymx=-18.35

gbr_ft_cropped <- st_crop(x1, xmin = xmn, ymin = ymn, xmax = xmx, ymax = ymx) #%>% filter(LEVEL_1=="Island")
gbr_coast_cropped <- st_crop(gbr_coast,xmin = xmn, ymin = ymn, xmax = xmx, ymax = ymx)

projcrs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

sites_sf <- st_as_sf(sites_data,coords = c("lon", "lat"),crs = projcrs)

sf_oz <- ozmap("states")
```

```{r}
ggplot() + 
  geom_sf(data = gbr_ft_cropped %>% filter(LEVEL_1=="Island")) +
  geom_sf(data = gbr_coast_cropped) +
  geom_sf(data = sites_sf,color="blue") + 
  ggrepel::geom_text_repel(data = sites_sf, mapping = aes(label=name, geometry=geometry), stat = "sf_coordinates",size=3) + 
  xlab("") + ylab("")
```

From this plot we can devise a rough north-south ordering for our locations while also ordering samples within Magnetic Island clockwise around the Island. We use this when plotting admixture results

```{r, echo=TRUE}
location_order <- 1:15
names(location_order) <- c("BRR","EP","WP","LPB","SO","JB","KR","HaR","HFB","WB","MB","HB","GB","PB","MR")
```

Create a stacked barplot to show admixture proportions

```{r}
run12_meta %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x = ID, y = p)) +
  scale_fill_manual(values = cluster_colors) +
  geom_col(aes(fill=cluster),position="stack", stat = "identity",width = 1,linewidth=0) +
  facet_grid(cols=vars(reorder(pop,pop_order)),scales = "free", space = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom") + 
  theme(strip.text = element_text(size=6)) +
  xlab("Sample ID") + ylab("Admixture Proportion")
```

```{r}
ggsave("figures/admixture_all.png",width = 3200,height = 800,units = "px")
```

For a more detailed investigation into single bays, we generated individual plots for each location and save into `figures`.

```{r, eval=FALSE}
# Assign a certain colour to the clusters, so that cluster1 is consistently blue and cluster2 orange
cluster_colors <- c("orange", "blue")

admix_plot_pop <- function(popdata,suffix="12",cluster_colors=c("orange", "blue")){
  # popdata will be a named vector with length 1.  We can get the name code and use this to look up the title
  code <- names(popdata)
  title <- location_names[code]
  
  p <- ggplot(popdata[[1]], aes(x = ID, y = p, fill = cluster)) +
  geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
   theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle(title)
  
  path <- paste("figures/",code,suffix,".png",sep="")
  ggsave(path,p,width = 1600,height = 800, units = "px")
  p
}

# Use lmap to iterate over all pops and run the plotting function on each
admix_plots_by_pop <- run12_meta %>% 
  split(run12_meta$pop) %>% 
  lmap(admix_plot_pop)
```

This is the dominant population structure evident in our dataset and is easily visible in a PCoA. Note that there is some additional variation along PC2, however this comprises a much lower proportion of genetic variation than PC1.

```{r}
if ( file.exists("cache/pcoa_ak_filtered_nr.rdata")){
  load("cache/pcoa_ak_filtered_nr.rdata")
} else {
  # Perform the PCoA
  pcoa_ak_filtered_nr <- gl.pcoa(ak.filtered.nr)
  # Cache the result because this is extremely slow
  save(pcoa_ak_filtered_nr,file = "cache/pcoa_ak_filtered_nr.rdata")
}

pcoa_var <- (pcoa_ak_filtered_nr$eig/sum(pcoa_ak_filtered_nr$eig))[1:2]

pcoa_ak_filtered_nr_meta <- pcoa_ak_filtered_nr$scores %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  left_join(metadata, by="ID") 
```

```{r}
pcoa_ak_filtered_nr_meta %>% 
  mutate(major_lineage = ifelse(pop %in% maggie_sites,"Magnetic Island","Surrounding Reefs")) %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=major_lineage)) + 
  xlab(paste("PC1 (",round(pcoa_var[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var[2]*100,1)," %)",sep=""))
```

# Removing migrants and highly admixed individuals

To investigate potential fine-scale population structure at Magnetic Island and adjacent reefs we removed migrants and highly admixed individuals (recent hybrids). Removal of these individuals was performed for all further population structure and genetic diversity analysis for the following reasons.

1.  Migrants (genetic origin different from geographic sampling location) were removed because the geographic location of parental origin of these individuals is unknown. Their alleles reflect population genetic processes at their parent reef and not the reef they were sampled on.

2.  Highly admixed individuals representing likely recent hybrids were removed. Although these individuals might represent the initial stages of gene flow between populations there is a strong possibility that such initial stages of gene flow (F1 hybrids and early backcrosses) are subject to processes that play a much reduced-role in longer-term gene flow. In particular such early hybrids may have reduced fertility, so including them in our analyses could lead to over-estimates of gene flow between populations.

3.  The overall number of highly admixed individuals is too small to justify a separate analysis of hybridisation processes in its own right.

4.  Fine-scale population structure within Magnetic Island and within adjacent reefs is likely to reflect recent processes (last 10kya) whereas the genetic distinction between Magnetic Island and adjacent reefs has been shown to be ancient (\>500kya). Identification of the (likely more subtle) genetic signatures of these recent processes could be swamped by false signal arising from inclusion of migrants and/or recent hybrids.

Based on the admixture results with K=2 and geographical location we consider "HFB","WB","MB","HB","GB","PB","MR" to represent Magnetic Island all other locations to be adjacent reefs.

We therefore flag for removal, any individuals within the Magnetic Island locations (ie "HFB","WB","MB","HB","GB","PB","MR" ) where the proportion of the opposing cluster (ie C1) is more than 5%, and vice-versa for adjacent reef locations.

Note that with this threshold of 5% we are only removing a total of 4 recent hybrids (admix\~0.3-0.5) spread across 4 locations, three recent backcrosses (admix 0.15-0.3) and one higher level backcross (0.05-0.15). These individuals are found at 5 reefs, of which only one HB is on Magnetic Island. Interestingly Havannah Reef, the closest to Magnetic Island, contained the largest number of putative recent hybrids (3) of any reef.

A summary table of putative backcross types based on admixture proportions is shown below (F1 (0.3-0.5), F2 (0.15=0.3), F2+ (0.05-0.15)

```{r}
run12_meta %>% 
  group_by(ID,pop) %>% 
  summarise(backcross_type = case_when(
    between(min(p),0,0.05) ~ "PB",
    between(min(p),0.05,0.15) ~ "F2+",
    between(min(p),0.15,0.3) ~ "F2",
    between(min(p),0.3,0.5) ~ "F1",
  )) %>% 
  ungroup() %>% 
  filter(backcross_type!="PB") %>% 
  group_by(pop,backcross_type) %>% 
  summarise(n=n())
```

```{r}
maggie_admixed_and_migrants <- run12_meta %>% 
  filter(pop %in% maggie_sites) %>% 
  filter(cluster=="C1") %>% 
  filter(p>0.05) %>% 
  pull(ID)

# And do the same for adjacent reef populations
adjacent_admixed_and_migrants <- run12_meta %>% 
  filter(!(pop %in% maggie_sites)) %>% 
  filter(cluster=="C2") %>% 
  filter(p>0.05) %>% 
  pull(ID)
```

In addition to these 7 highly admixed individuals we found a further 17 migrants. In total we removed 9 individuals from adjacent reefs and 14 from Magnetic Island.

The resulting filtered dataset (with these individuals removed) is saved as `ak.pop.nm`.

```{r}
if ( file.exists("cache/ak.pop.nm.rdata")){
  load("cache/ak.pop.nm.rdata")
} else {

  ak.pop.nm <- gl.drop.ind(ak.filtered.nr, ind.list=c(maggie_admixed_and_migrants,adjacent_admixed_and_migrants))

  # Filter for monomorphs
  ak.pop.nm <- gl.filter.monomorphs(ak.pop.nm)

  # Recalculate locus metrics
  ak.pop.nm <- gl.recalc.metrics(ak.pop.nm)

  ## Remove missing loci
  #A DArT dataset will not have individuals for which the calls are scored all as missing (NA) across all loci, but such individuals may   #sneak in to the dataset when loci are deleted. Retaining individual or loci with all NAs can cause issues for several functions

  ak.pop.nm <- gl.filter.allna(ak.pop.nm)

  nInd(ak.pop.nm)
  nLoc(ak.pop.nm)

  save(ak.pop.nm, file="cache/ak.pop.nm.rdata")
}
```

# Fine-scale population structure

To examine patterns of population structure in further detail we begin by calculating pairwise Fst between all locations.

```{r}
# Calculate pairwise Fst values between populations
if ( file.exists("cache/fst.pop.rdata")){
  load("cache/fst.pop.rdata")
} else {
  fst.pop <- gl.fst.pop(ak.pop.nm, nboots=100, percent=95, nclusters = min(4,parallel::detectCores()))
  save(fst.pop,file = "cache/fst.pop.rdata")
}
```

This very clearly recapitulates the strong structure already evident at the broadscale between Magnetic Island and adjacent populations as evidenced in this heatmap

```{r}
fst.matrix <- as.matrix(as.dist(fst.pop$Fsts))
heatmap(fst.matrix)
```

Quantitatively we see that Fst values between Magnetic Island and adjacent are much higher (\~0.3) than between sites within the same broad-scale population grouping (ie within Maggie or within adjacent). We can also see that Fst values between maggie bays are broadly spread between -0.005 and 0.01 whereas Fst values between adjacent sites seem to show two groupings. This suggests that there might be some fine-scale structure within the adjacent sites.

```{r}
fst_comparisons <- fst.matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("pop1") %>% 
  pivot_longer(-1, names_to = "pop2", values_to = "Fst") %>% 
  mutate(cluster1 = case_when(
    pop1 %in% maggie_sites ~ "maggie",
    .default = "adjacent"
  )) %>% 
  mutate(cluster2 = case_when(
    pop2 %in% maggie_sites ~ "maggie",
    .default = "adjacent"
  )) %>% 
  mutate(fst_type = case_when(
    (cluster1 == cluster2) & (cluster1 == "maggie") ~ "within_maggie",
    (cluster1 == cluster2) & (cluster1 == "adjacent") ~ "within_adjacent",
    (cluster1 != cluster2) ~ "between",    
  )) %>% 
  mutate(comp_type = str_extract(fst_type,"([^_]*)"))


ggplot(fst_comparisons,aes(x=fst_type,y=Fst)) + 
  geom_boxplot(outliers = FALSE) +
  geom_jitter(aes(color=fst_type)) + 
  facet_wrap(~comp_type, scales = "free")
```

A closer look at those elevated Fst comparisons within adjacent reveals that they all caused by slight divergence of WP from other adjacent sites.

```{r}
fst_comparisons %>% 
  filter(fst_type=="within_adjacent") %>% 
  filter(Fst>0.005) 
```

## Fine Scale Structure within adjacent and within Magnetic Island

First we subset data for each location, including only relevant sites.

```{r}
if ( file.exists("cache/ak.pop.nm.adj.rdata")){
  load("cache/ak.pop.nm.adj.rdata")
} else {
  # Keep only subpopulations from Adjacent sites for fine structure analysis
  ak.pop.nm.adj <- gl.keep.pop(ak.pop.nm, pop.list = as.list(setdiff(location_codes,maggie_sites)))
  save(ak.pop.nm.adj, file="cache/ak.pop.nm.adj.rdata")
}

if ( file.exists("cache/ak.pop.nm.mag.rdata")){
  load("cache/ak.pop.nm.mag.rdata")
} else {
  # Keep only subpopulations from Magnetic Island for fine structure analysis
  ak.pop.nm.mag <- gl.keep.pop(ak.pop.nm, pop.list = list("GB", "HB", "HFB", "MB", "MR", "PB", "WB"))
  save(ak.pop.nm.mag, file="cache/ak.pop.nm.mag.rdata")
}
```

Then we convert both subsetted datasets to structure format to run externally with structure program. For both datasets we performed 20 replicate STRUCTURE runs, each with 20000 steps and 10000 burnin. Generating 20 replicates allows us to calculate the $\Delta K$ statistic of Evanno et al.

```{r, eval=FALSE}
# Convert to structure format to run externally with structure program
gl2structure(ak.pop.nm.mag, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.pop.mag.nm.str", outpath = "structure", verbose = NULL)

gl2structure(ak.pop.nm.adj, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.pop.adj.nm.str", outpath = "structure", verbose = NULL)
```

The plot of $\Delta K$ for Magnetic Island reveals a clear preference for K=3 at that location but showed a more complex picture for the adjacent locations, where we see elevated values for K=2 and K=5. Noting that in this method it is not possible to include K=1 in the test our decision on the choice of $K$ must also be informed by other factors such as PCA.

```{r}
read_deltak_data <- function(subset="adj"){
  like_filename <- paste("k_vs_lik_",subset,".txt",sep = "")
  llpaths <- list.files("structure",like_filename,recursive = TRUE,full.names = TRUE)
  reps <- llpaths %>% str_extract("([0-9]+)")

  lldata <- map2(llpaths,reps,read_loglik) %>% 
    list_rbind()

  lldata %>% 
    group_by(k) %>% 
    summarise(deltaK = mean(abs(deltaL))/var(logLik)) %>% 
    add_column(subset=subset)
}

deltak_data <- map(c("adj","mag"),read_deltak_data) %>% 
  list_rbind()

ggplot(deltak_data %>% na.omit(),aes(x=k,y=deltaK)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~subset) +
  xlab("Number of clusters, K") + 
  ylab(expression(paste(Delta,"K")))
```

#### Fine structure within adjacent

For the adjacent population we begin by examining the admixture result with $K=2$ since this is the lowest $K$ with an elevated $\Delta K$ value. This reveals a model driven by just two individuals from WP representing the blue cluster, while all other individuals are putative F1 hybrids between an unknown population and these WP individuals. This is not biologically realistic and suggests that the two WP individuals are outliers.

```{r, warning=FALSE}
# Reading admixture results for Adjacent and K=2

samples_adj <- read_table("structure/ak.pop.adj.nm.samples.txt",col_names = c("ID"))

run_adj_k2 <- read_table("structure/adj.1/ak.pop.adj.nm.k2.out.anc.txt",col_names = c("X1", "Label","(%Miss)","C1","C2"),skip = 2) %>% 
  cbind(samples_adj) %>% 
  select(ID,starts_with("C")) %>% # Select only useful columns
  pivot_longer(-ID,names_to = "cluster",values_to = "p") %>% 
  left_join(metadata,by="ID")

run_adj_k2 %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x = ID, y = p)) +
  geom_col(aes(fill=cluster),position="stack", stat = "identity",width = 1,linewidth=0) +
  facet_grid(cols=vars(reorder(pop,pop_order)),scales = "free", space = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom") + 
  theme(strip.text = element_text(size=6)) +
  xlab("Sample ID") + ylab("Admixture Proportion")
```

This can also be seen in a PCA which shows the two WP individuals as outliers.

```{r}
if (file.exists("cache/pcoa_ak_filtered_nr_adj.rdata")){
  load("cache/pcoa_ak_filtered_nr_adj.rdata")
} else {
  # Note that despite its name gl.pcoa performs a PCA not a PCoA in this case
  pcoa_ak_filtered_nr_adj <- gl.pcoa(ak.pop.nm.adj, plot.out = FALSE)
  save(pcoa_ak_filtered_nr_adj,file = "cache/pcoa_ak_filtered_nr_adj.rdata")
}
```

```{r}
outlier_ids_adj <- run_adj_k2 %>% 
  filter(p>0.9) %>% pull(ID)

pcoa_var_adj <- (pcoa_ak_filtered_nr_adj$eig/sum(pcoa_ak_filtered_nr_adj$eig))[1:2]

pcoa_ak_filtered_nr_meta_adj <-  pcoa_ak_filtered_nr_adj$scores %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  left_join(metadata,by="ID") 

pcoa_ak_filtered_nr_meta_adj %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=pop)) + 
  geom_label_repel(data = (pcoa_ak_filtered_nr_meta_adj %>% filter(ID %in% outlier_ids_adj)),aes(label=ID)) +
  xlab(paste("PC1 (",round(pcoa_var_adj[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var_adj[2]*100,1)," %)",sep=""))
```

These individuals might represent members of a third major population cluster for this species. According to Matias et al our adjacent cluster is likely to be XX based on location but individials from other clusters (XX) may be present.

Removing these individuals and rerunning PCA reveals a complete absence of additional structure

```{r}

```

```{r}
ggsave("figures/admixture_all_mag.png",width = 3200,height = 800,units = "px")
```

### Fine Scale Structure at Magnetic Island

### Load csv

```{r, warning=FALSE}
samples10 <- read_table("structure/ak.pop.mag.nm.samples.txt",col_names = c("ID"))

run10 <- read_table("structure/ak.pop.mag.nm.k3.out.anc.txt",col_names = c("X1", "Label","(%Miss)","C1","C2","C3"),skip = 2) %>% 
  cbind(samples10) %>% 
  select(ID,starts_with("C")) %>% # Select only useful columns
  pivot_longer(-ID,names_to = "cluster",values_to = "p") %>% 
  left_join(metadata,by="ID")
```

For a first glimpse, create a barplot that shows the bays on the x-axis, admixture proportion on y-axis and all clusters stacked.

```{r}
run10 %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x = ID, y = p)) +
  geom_col(aes(fill=cluster),position="stack", stat = "identity",width = 1,linewidth=0) +
  facet_grid(cols=vars(reorder(pop,pop_order)),scales = "free", space = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom") + 
  theme(strip.text = element_text(size=6)) +
  xlab("Sample ID") + ylab("Admixture Proportion")
```

```{r}
ggsave("figures/admixture_all_mag.png",width = 3200,height = 800,units = "px")
```

Plots for individual bays are saved into the figures folder

```{r, eval=FALSE}
# Use lmap to iterate over all pops and run the plotting function on each
admix_plots_by_pop10 <- run10 %>% 
  split(run10$pop) %>% 
  lmap(admix_plot_pop,"10",cluster_colors = c(cluster_colors,"cyan","purple"))
```

Lets use a PCA now to try and understand our result of 3 clusters and see if that matches what we see using a different visualisation technique.

```{r}
load("cache/ak.pop.nm.mag.rdata")

if (file.exists("cache/pcoa_ak_filtered_nr_mag.rdata")){
  load("cache/pcoa_ak_filtered_nr_mag.rdata")
} else {
  pcoa_ak_filtered_nr_mag <- gl.pcoa(ak.pop.nm.mag, plot.out = FALSE)
  save(pcoa_ak_filtered_nr_mag,file = "cache/pcoa_ak_filtered_nr_mag.rdata")
}

dominant_clusters_run10 <- run10 %>% 
  group_by(ID) %>% 
  slice_max(p) %>% 
  select(ID,cluster,p)


pcoa_var_mag <- (pcoa_ak_filtered_nr_mag$eig/sum(pcoa_ak_filtered_nr_mag$eig))[1:2]

pcoa_ak_filtered_nr_meta_mag <-  pcoa_ak_filtered_nr_mag$scores %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  left_join(metadata,by="ID") %>% 
  left_join(dominant_clusters_run10,by="ID")
```

When samples are annotated with their dominant cluster we can see that the three apparent groupings in PCA do indeed correspond with groupings identified by structure.

```{r}
pcoa_ak_filtered_nr_meta_mag %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=cluster)) + 
  xlab(paste("PC1 (",round(pcoa_var_mag[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var_mag[2]*100,1)," %)",sep=""))
```

Now plotting by location we can see that the largest cluster (C1) is comprised of samples from all bays, however, the two smaller clusters (C3, and C2) are primarily made up of individuals from Picnic Bay and Geoffrey Bay. Beyond this there is no apparent geographic pattern to genetic structure within the Magnetic Island population.

```{r}
pcoa_ak_filtered_nr_meta_mag %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x=PC1,y=PC2)) + geom_point(aes(color=reorder(pop,pop_order))) + 
  xlab(paste("PC1 (",round(pcoa_var_mag[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var_mag[2]*100,1)," %)",sep=""))
```
