---
title: "3 Population structure"
author: "Sandra Erdmann"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: inline
execute:
  freeze: auto
---

## Load Libraries

```{r}
library(ggplot2)      # to plot results
library(ggpubr)       # to use ggarrange and other plotting functions
library(gridExtra)    # to use grid.arrange
library(cowplot)      # to use plot_grid and ggdraw
library(tidyr)        # general functions
library(dartRverse)
```





```{r}
load("cache/ak.pop.rdata")
```




## Convert to structure

Convert ak.pop to structure format to run externally with structure program

```{r}
gl2structure(ak.pop, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.pop.str", outpath = "structure", verbose = NULL)
```

```{r}
nInd(ak.pop)
nLoc(ak.pop)
```

TODO: Add details of how structure was run here

## Admixture plot for ak.pop

### Load csv

Added the location to csv, so it is possible to group by location for plots.

```{r}
# Load csv with data from run # 12 in the structure analysis
run12 <-  read.csv("ak.pop_run12.csv")
```

```{r}
head(run12)
```

### Data wrangling

For the plots to show both clusters the matrix needs to be amended.

```{r}
# Create 2 rows per ID for each cluster, assign the fst value and arrange by ID
run12<-run12 %>% gather(key="cluster", value = "fst", c(-ID, -loc)) %>% arrange(ID)
```

```{r}
head(run12)
```

For a first glimpse, create a barplot that shows the bays on the x-axis, fst on y-axis and all 4 clusters stacked.

```{r}
strrun12 <- ggplot(run12, aes(x = ID, y = fst, fill=cluster)) +
   geom_bar(position="stack", stat = "identity", width=0.2)
strrun12
```

For a more detailed investigation into single bays, select the location from the dataset and make subsets for all locations ("HFB", "WB", "MB", "HB", "GB", "PB", "MR", "HaR", "SO", "LPB", "WP", "EP", "BRR", "JB", "KR")

```{r}
HFBrun12 <-run12[grep("^HFB", run12$loc), ] 
HFBrun12
```

```{r}
WBrun12<-run12[grep("^WB", run12$ID), ]
WBrun12
```

```{r}
MBrun12<-run12[grep("^MB", run12$ID), ]   
MBrun12
```

```{r}
HBrun12<-run12[grep("^HB", run12$ID), ]
HBrun12
```

```{r}
GBrun12<-run12[grep("^GB", run12$ID), ]
GBrun12
```

```{r}
PBrun12<-run12[grep("^PB", run12$ID), ]
PBrun12
```

```{r}
MRrun12<-run12[grep("^MR", run12$ID), ]
MRrun12
```

```{r}
HaRrun12<-run12[grep("^HaR", run12$ID), ]   
HaRrun12
```

```{r}
SOrun12 <-run12[grep("^SO", run12$loc), ] 
SOrun12
```

```{r}
LPBrun12 <-run12[grep("^LPB", run12$loc), ]  
LPBrun12
```

```{r}
WPrun12 <-run12[grep("^WP", run12$loc), ] 
WPrun12
```

```{r}
EPrun12 <-run12[grep("^EP", run12$loc), ] 
EPrun12
```

```{r}
BRRrun12<-run12[grep("^BRR", run12$ID), ]     #  samples
BRRrun12
```

```{r}
JBrun12<-run12[grep("^JB", run12$ID), ]  
JBrun12
```

```{r}
KRrun12<-run12[grep("^KR", run12$ID), ] 
KRrun12
```

### Plot

```{r}
# Assign a certain colour to the clusters, so that cluster1 is consistently blue and cluster2 orange
cluster_colors <- c("orange", "blue")
```

Make a plot for all locations ("HFB", "WB", "MB", "HB", "GB", "PB", "MR", "HaR", "SO", "LPB", "WP", "EP", "BRR", "JB", "KR").

```{r}
HFB12 <- ggplot(HFBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
   theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
  ggtitle("Huntingfield Bay")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/HFB12.png",
width=800, height=400)

HFB12
```

```{r}
WB12 <- ggplot(WBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
  ggtitle("Wilson Bay")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/WB12.png",
width=800, height=400)

WB12
```

```{r}
MB12 <- ggplot(MBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
   theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Maud Bay")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/MB12.png",
width=800, height=400)

MB12
```

```{r}
HB12 <- ggplot(HBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
    theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
  ggtitle("Horseshoe Bay")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/HB12.png",
width=800, height=400)

HB12
```

```{r}
GB12 <- ggplot(GBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Geoffrey Bay")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/GB12.png",
width=800, height=400)

GB12
```

```{r}
PB12 <- ggplot(PBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Picnic Bay")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/PB12.png",
width=800, height=400)

PB12
```

```{r}
MR12 <- ggplot(MRrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Middle Reef")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/MR12.png",
width=800, height=400)

MR12
```

```{r}
HaR12 <- ggplot(HaRrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors)  +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Havannah Reef")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/HaR12.png",
width=800, height=400)

HaR12
```

```{r}
SO12 <- ggplot(SOrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("South Orpheus")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/SO12.png",
width=800, height=400)

SO12
```

```{r}
LPB12 <- ggplot(LPBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Little Pioneer Bay")


png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/LPB12.png",
width=800, height=400)

LPB12
```

```{r}
WP12 <- ggplot(WPrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
      ggtitle("West Pelorus")


png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/WP12.png",
width=800, height=400)

WP12
```

```{r}
EP12 <- ggplot(EPrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
     ggtitle("East Pelorus")


png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/EP12.png",
width=800, height=400)

EP12
```

```{r}
BRR12 <- ggplot(BRRrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Bramble Reef")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/BRR12.png",
width=800, height=400)

BRR12
```

```{r}
JB12 <- ggplot(JBrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("John Brewer Reef")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/JB12.png",
width=800, height=400)

JB12
```

```{r}
KR12 <- ggplot(KRrun12, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Keeper Reef")

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/KR12.png",
width=800, height=400)

KR12
```

Combine all structure plots into one plot.

```{r}
# Combine all Magnetic Island subpopulations
# Remove both x and y axis elements (including labels, ticks, and lines) from all but the first column and combine all plots
strmag12_combined <- ggarrange(
  HFB12 + theme(axis.title.x = element_blank(),
                axis.text.x = element_blank(),
                axis.ticks.x = element_blank(),
                axis.title.y = element_blank(),
                axis.text.y = element_blank(),
                axis.ticks.y = element_blank()),  # Remove both axes
  WB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  MB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  HB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  GB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  PB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  MR12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  ncol = 2, nrow = 4,
  widths = c(1, 1),
  heights = c(1, 1),
  legend = "none"
)

# Add a common y-axis label using cowplot::plot_grid()
strmag12_labeled <- plot_grid(
  ggdraw() + 
    draw_label("Admixture proportions", angle = 90, size = 15, vjust = 0.5),
  strmag12_combined,
  ncol = 2,
  rel_widths = c(0.05, 1)  # 5% for label, 95% for plot
)

# Save the plot
ggsave("C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures//strmag12.pdf", plot = strmag12_labeled, width = 8, height = 10)
```

```{r}
# Combine all adjacent reef populations
# Remove both x and y axis elements (including labels, ticks, and lines) from all but the first column and combine all plots
stradj12_combined <- ggarrange(
  HaR12 + theme(axis.title.x = element_blank(),
                axis.text.x = element_blank(),
                axis.ticks.x = element_blank(),
                axis.title.y = element_blank(),
                axis.text.y = element_blank(),
                axis.ticks.y = element_blank()),  # Remove both axes
  SO12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  LPB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  WP12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  EP12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  BRR12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  JB12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  KR12 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  ncol = 2, nrow = 4,
  widths = c(1, 1),
  heights = c(1, 1),
  legend = "none"
)

# Add a common y-axis label using cowplot::plot_grid()
stradj12_labeled <- plot_grid(
  ggdraw() + 
    draw_label("Admixture proportions", angle = 90, size = 15, vjust = 0.5),
  stradj12_combined,
  ncol = 2,
  rel_widths = c(0.05, 1)  # 5% for label, 95% for plot
)

# Save the plot
ggsave("C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures//stradj12.pdf", plot = stradj12_labeled, width = 8, height = 10)
```

# Fine-scale population structure for Magnetic Island

## Remove migrants

For fine scale structure at Magnetic Island migrants from adjacent reefs will be removed to create a subset only for the Magnetic Island population called ak.pop.nm.mag

```{r}
# Get names for all individuals
indNames(ak.pop)
```

The following migrants, which have allele frequency differences \>0.5, have been identified with structure plots: HB: 1, 2, 14, 20, 21, 2021_11 WB: 8, 25, 37 HFB: 4, 6, 9, 10, 22 HaR: 7, 10, 12, 24, 43

```{r}
# Remove 19 migrants with a proportion >5%
ak.pop.nm <- gl.drop.ind(ak.pop, ind.list=c("HB_1", "HB_2", "HB_27/11/21_11", "HB_27/11/21_14", "HB_27/11/21_20", "HB_27/11/21_21", "WB_03/04/2022_8", "WB_03/04/2022_25", "WB_03/04/2022_37", "HFB_4", "HFB_6", "HFB_9", "HFB_10", "HFB_22", "HaR_06/05/2022_7", "HaR_06/05/2022_10", "HaR_06/05/2022_12", "HaR_06/05/2022_24", "HaR_06/05/2022_43"))
```

```{r}
# Filter for monomorphs
ak.pop.nm <- gl.filter.monomorphs(ak.pop.nm)
```

```{r}
# Recalculate locus metrics
ak.pop.nm <- gl.recalc.metrics(ak.pop.nm)
```

## Remove missing loci

A DArT dataset will not have individuals for which the calls are scored all as missing (NA) across all loci, but such individuals may sneak in to the dataset when loci are deleted. Retaining individual or loci with all NAs can cause issues for several functions

```{r}
# Filter for missing loci
ak.pop.nm <- gl.filter.allna(ak.pop.nm)
```

```{r}
nInd(ak.pop.nm)
nLoc(ak.pop.nm)
```

```{r}
gl.smearplot(ak.pop.nm)
```

```{r}
save(ak.pop.nm, file="ak.pop.nm.rdata")
```

## Keep only Magnetic Island subpopulations

```{r}
indNames(ak.pop.nm)
```

```{r}
# Keep only subpopulations from Magnetic Island for fine structure analysis
ak.pop.mag <- gl.keep.pop(ak.pop.nm, pop.list = list("GB", "HB", "HFB", "MB", "MR", "PB", "WB"))
```

```{r}
save(ak.pop.mag, file="ak.pop.mag.rdata")
```

```{r}
nInd(ak.pop.mag)
nLoc(ak.pop.mag)
```

## Convert to structure

Convert ak.pop.mag to structure format to run externally with structure program

```{r}
gl2structure(ak.pop.mag, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.pop.mag.str", outpath = NULL, verbose = NULL)
```

```{r}
nInd(ak.pop.mag)
nLoc(ak.pop.mag)
```

## Admixture plot for ak.pop.nm.mag

### Load csv

Added the location to csv, so it is possible to group by location for plots.

```{r}
# Load csv with data from run # 12 in the structure analysis
run10 <-  read.csv("ak.pop.nm.mag_run10.csv")
```

```{r}
head(run10)
```

### Data wrangling

For the plots to show both clusters the matrix needs to be amended.

```{r}
# Create 2 rows per ID for each cluster, assign the fst value and arrange by ID
run10<-run10 %>% gather(key="cluster", value = "fst", c(-ID, -loc)) %>% arrange(ID)
```

```{r}
head(run10)
```

For a first glimpse, create a barplot that shows the bays on the x-axis, fst on y-axis and all 4 clusters stacked.

```{r}
strrun10 <- ggplot(run10, aes(x = ID, y = fst, fill=cluster)) +
   geom_bar(position="stack", stat = "identity", width=0.2)
strrun10
```

For a more detailed investigation into single bays, select the location from the dataset and make subsets for all locations ("HFB", "WB", "MB", "HB", "GB", "PB", "MR").

```{r}
HFBrun10 <-run10[grep("^HFB", run10$loc), ] 
HFBrun10
```

```{r}
WBrun10<-run10[grep("^WB", run10$ID), ]
WBrun10
```

```{r}
MBrun10<-run10[grep("^MB", run10$ID), ]   
MBrun10
```

```{r}
HBrun10<-run10[grep("^HB", run10$ID), ]
HBrun10
```

```{r}
GBrun10<-run10[grep("^GB", run10$ID), ]
GBrun10
```

```{r}
PBrun10<-run10[grep("^PB", run10$ID), ]
PBrun10
```

```{r}
MRrun10<-run10[grep("^MR", run10$ID), ]
MRrun10
```

### Plot

```{r}
# Assign a certain colour to the clusters, so that cluster1 is consistently blue and cluster2 orange
cluster_colors <- c("blue", "orange")
```

Make a plot for all locations ("HFB", "WB", "MB", "HB", "GB", "PB", "MR", "HaR", "SO", "LPB", "WP", "EP", "BRR", "JB", "KR").

```{r}
HFB10 <- ggplot(HFBrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
  ggtitle("Huntingfield Bay") +
   theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/HFB10.png",
width=800, height=400)

HFB10
```

```{r}
WB10 <- ggplot(WBrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
  ggtitle("Wilson Bay") +
    theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/WB10.png",
width=800, height=400)

WB10
```

```{r}
MB10 <- ggplot(MBrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Maud Bay") +
   theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/MB10.png",
width=800, height=400)

MB10
```

```{r}
HB10 <- ggplot(HBrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
  ggtitle("Horseshoe Bay") +
    theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/HB10.png",
width=800, height=400)
HB10
```

```{r}
GB10 <- ggplot(GBrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Geoffrey Bay") +
  theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/GB10.png",
width=800, height=400)
GB10
```

```{r}
PB10 <- ggplot(PBrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Picnic Bay") +
  theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/PB10.png",
width=800, height=400)

PB10
```

```{r}
MR10 <- ggplot(MRrun10, aes(x = ID, y = fst, fill = cluster)) +
   geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle("Middle Reef") +
  theme_classic()

png(file="C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/MR10.png",
width=800, height=400)
MR10
```

Combine all structure plots into one plot.

```{r}
# Combine all Magnetic Island subpopulations
# Remove both x and y axis elements (including labels, ticks, and lines) from all but the first column and combine all plots
strmaggie10_combined <- ggarrange(
  HFB10 + theme(axis.title.x = element_blank(),
                axis.text.x = element_blank(),
                axis.ticks.x = element_blank(),
                axis.title.y = element_blank(),
                axis.text.y = element_blank(),
                axis.ticks.y = element_blank()),  # Remove both axes
  WB10 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  MB10 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  HB10 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  GB10 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  PB10 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  MR10 + theme(axis.title.x = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank(),
               axis.title.y = element_blank(),
               axis.text.y = element_blank(),
               axis.ticks.y = element_blank()),  # Remove both axes
  ncol = 2, nrow = 4,
  widths = c(1, 1),
  heights = c(1, 1),
  legend = "none"
)

# Add a common y-axis label using cowplot::plot_grid()
strmaggie10_labeled <- plot_grid(
  ggdraw() + 
    draw_label("Admixture proportions", angle = 90, size = 15, vjust = 0.5),
  strmaggie10_combined,
  ncol = 2,
  rel_widths = c(0.05, 1)  # 5% for label, 95% for plot
)

# Save the plot
ggsave("C:/Users/sandr/Documents/Chapter 1 - Github Script/Figures/strmaggie10.pdf", plot = strmaggie10_labeled, width = 8, height = 10)
```
