---
title: "3 Population structure"
author: "Sandra Erdmann"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: inline
execute:
  freeze: auto
---

## Load Libraries

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)

library(tidyverse)
library(dartRverse)
library(dartR.base)
library(ggpubr)
library(sf)
library(ozmaps)
source("constants.R")

options(readr.show_col_types = FALSE)
```

```{r}
load("cache/ak.filtered.nr.rdata")
```

## Convert to structure

Convert `ak.filtered.nr` to structure format to run externally with structure program

```{r, eval=FALSE}
gl2structure(ak.filtered.nr, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.filtered.nr.str", outpath = "structure", verbose = NULL)
```

```{r}
nInd(ak.filtered.nr)
nLoc(ak.filtered.nr)
```

Structure was then run for K = 1, 2, 3 and 4. Each run used 20000 MCMC iterations with 10000 additional as burn-in. Full parameter files can be found in `structure`

Determining the appropriate value of K is tricky. The structure manual recommends care is taken, and that the lowest biologically sensible value is used. A more objective method of [Evanno 2005](https://app.paperpile.com/view/plain/?id=8d8f1080-2416-4bb5-97a5-504b82a919d7) involves taking the second derivative of the log Likelihood. This method suggests that K=2 is optimal.

```{r}
llpaths <- list.files("structure","k_vs_lik.txt",recursive = TRUE,full.names = TRUE)
reps <- llpaths %>% str_extract("([0-9]+)")

read_loglik <- function(path,n){
  read_table(path,col_names = c("k","logLik"))  %>%  
    mutate(Ldash = logLik - lag(logLik,1)) %>% 
    mutate(deltaL = lead(Ldash) - Ldash) %>% 
    add_column(rep=n)
}


lldata <- map2(llpaths,reps,read_loglik) %>% 
  list_rbind()

lldata %>% 
  group_by(k) %>% 
  summarise(deltaK = mean(abs(deltaL))/var(logLik)) %>% 
  ggplot(aes(x=k,y=deltaK)) + 
    geom_point() + geom_line() + xlab("Number of clusters, K") + ylab("L''")
```

## Admixture plot for K=2

Read assignment probabilities for individuals from the run with K=2.

```{r, warning=FALSE}
# Load csv with data from run # 12 in the structure analysis
run12_samples <- read_table("structure/ak.filtered.nr.samples.txt",col_names = c("ID"))

run12 <-  read_table("structure/ak.structure.nr.k2.out.anc.txt",col_names = c("X1", "Label","(%Miss)","C1","C2"),skip = 2) %>% 
  cbind(run12_samples) %>% 
  select(ID,C1,C2) %>% # Select only useful columns
  pivot_longer(-ID,names_to = "cluster",values_to = "p") # Put assignment probabilities into a single column
```

Join structure data with sample metadata

```{r}
metadata <- read_csv("ind_metrics_SNP2.csv") %>% 
  mutate(ID = case_when(
    id=="20" ~ "WB_36",
    .default = id
  )) %>% 
  select(-id)
save(metadata,file = "cache/metadata.rdata")


run12_meta <- run12 %>% 
  left_join(metadata,by="ID")
```

In order to make sense of population structure it is useful to visualise sampling locations on a map

```{r}
sites_data <- run12_meta %>% 
  group_by(pop,lat,lon) %>% 
  summarise(n_samples = n()/2) %>% 
  mutate(name = location_names[pop])
```

```{r, include=FALSE}
gbr_coast <- st_read("maps/GBR_NESP-TWQ-5/BaseLayers/",layer = "GBR_coastlineONLY")
gbr_tc <- st_read("maps/GBR_NESP-TWQ-5/BaseLayers/",layer = "GBR_towncities")
gbr_features <- st_read("maps/GBR_NESP-TWQ-5/BaseLayers/",layer = "TS_AIMS_NESP_Torres_Strait_Features_V1b_with_GBR_Features")

sf_use_s2(FALSE) # the input geometries are bad, and wrong spherically
x1 <- st_make_valid(gbr_features)

xmn=146.0
ymn=-19.3
xmx=146.9
ymx=-18.35

gbr_ft_cropped <- st_crop(x1, xmin = xmn, ymin = ymn, xmax = xmx, ymax = ymx) #%>% filter(LEVEL_1=="Island")
gbr_coast_cropped <- st_crop(gbr_coast,xmin = xmn, ymin = ymn, xmax = xmx, ymax = ymx)

projcrs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

sites_sf <- st_as_sf(sites_data,coords = c("lon", "lat"),crs = projcrs)

sf_oz <- ozmap("states")
```

```{r}
ggplot() + 
  geom_sf(data = gbr_ft_cropped %>% filter(LEVEL_1=="Island")) +
  geom_sf(data = gbr_coast_cropped) +
  geom_sf(data = sites_sf,color="blue") + 
  ggrepel::geom_text_repel(data = sites_sf, mapping = aes(label=name, geometry=geometry), stat = "sf_coordinates",size=3) + 
  xlab("") + ylab("")
```

From this plot we can devise a rough north-south ordering for our locations. This will help when plotting admixture results

```{r}
location_order <- 1:15
names(location_order) <- c("BRR","EP","WP","LPB","SO","JB","KR","HaR","HFB","WB","MB","HB","GB","PB","MR")
```

Create a stacked barplot to show admixture proportions

```{r}
run12_meta %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x = ID, y = p)) +
  scale_fill_manual(values = cluster_colors) +
  geom_col(aes(fill=cluster),position="stack", stat = "identity",width = 1,linewidth=0) +
  facet_grid(cols=vars(reorder(pop,pop_order)),scales = "free", space = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom") + 
  theme(strip.text = element_text(size=6)) +
  xlab("Sample ID") + ylab("Admixture Proportion")
ggsave("figures/admixture_all.png",width = 3200,height = 800,units = "px")
```

For a more detailed investigation into single bays, we generate individual plots for each location and save into `figures`.

```{r}
# Assign a certain colour to the clusters, so that cluster1 is consistently blue and cluster2 orange
cluster_colors <- c("orange", "blue")

admix_plot_pop <- function(popdata,suffix="12",cluster_colors=c("orange", "blue")){
  # popdata will be a named vector with length 1.  We can get the name code and use this to look up the title
  code <- names(popdata)
  title <- location_names[code]
  
  p <- ggplot(popdata[[1]], aes(x = ID, y = p, fill = cluster)) +
  geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
   theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  font("x.text", size = 8) +
    labs(y = "admixture proportions") +
    ggtitle(title)
  
  path <- paste("figures/",code,suffix,".png",sep="")
  ggsave(path,p,width = 1600,height = 800, units = "px")
  p
}

# Use lmap to iterate over all pops and run the plotting function on each
admix_plots_by_pop <- run12_meta %>% 
  split(run12_meta$pop) %>% 
  lmap(admix_plot_pop)
```


This is the dominant population structure evident in our dataset and is easily visible in a PCoA. Note that there is some additional variation along PC2, however this comprises a much lower proportion of genetic variation than PC1.  

```{r}
# Perform the PCoA
pcoa_ak_filtered_nr <- gl.pcoa(ak.filtered.nr)

pcoa_var <- (pcoa_ak_filtered_nr$eig/sum(pcoa_ak_filtered_nr$eig))[1:2]

pcoa_ak_filtered_nr_meta <- pcoa_ak_filtered_nr$scores %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  left_join(metadata, by="ID") 
```


```{r}
pcoa_ak_filtered_nr_meta %>% 
  mutate(major_lineage = ifelse(pop %in% maggie_sites,"Magnetic Island","Surrounding Reefs")) %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=major_lineage)) + 
  xlab(paste("PC1 (",round(pcoa_var[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var[2]*100,1)," %)",sep=""))
```


# Fine-scale population structure for Magnetic Island

## Remove migrants and highly admixed individuals

For fine scale structure at Magnetic Island migrants and highly admixed individuals from adjacent reefs will be removed to create a subset only for the Magnetic Island population called ak.pop.nm.mag. This is because those individuals are likely to distort any estimates of location population structure (ie due to restricted gene flow) by introducing a different process (admixture with a highly diverged population).

On this basis we want to keep individuals within the Magnetic Island locations (ie "HFB","WB","MB","HB","GB","PB","MR" ) and where the proportion of the opposing cluster (ie C1) is more than 5%.  

Note that with this threshold of 5% we are only removing a total of 4 recent hybrids (admix~0.3-0.5) spread across 4 locations, three recent backcrosses (admix 0.15-0.3) and one higher level backcross (0.05-0.15). These individuals are not concentrated in any particular location.

```{r}
run12_meta %>% 
  group_by(ID,pop) %>% 
  summarise(backcross_type = case_when(
    between(min(p),0,0.05) ~ "PB",
    between(min(p),0.05,0.15) ~ "F2+",
    between(min(p),0.15,0.3) ~ "F2",
    between(min(p),0.3,0.5) ~ "F1",
  )) %>% 
  ungroup() %>% 
  group_by(pop,backcross_type) %>% 
  summarise(n=n()) %>% view()
```


```{r}
maggie_admixed_and_migrants <- run12_meta %>% 
  filter(pop %in% maggie_sites) %>% 
  filter(cluster=="C1") %>% 
  filter(p>0.05) %>% 
  pull(ID)

# And do the same for adjacent reef populations
adjacent_admixed_and_migrants <- run12_meta %>% 
  filter(!(pop %in% maggie_sites)) %>% 
  filter(cluster=="C2") %>% 
  filter(p>0.05) %>% 
  pull(ID)
```

```{r}
# Remove migrant and admixed individuals across the entire dataset
ak.pop.nm <- gl.drop.ind(ak.filtered.nr, ind.list=c(maggie_admixed_and_migrants,adjacent_admixed_and_migrants))
```

```{r}
# Filter for monomorphs
ak.pop.nm <- gl.filter.monomorphs(ak.pop.nm)
```

```{r}
# Recalculate locus metrics
ak.pop.nm <- gl.recalc.metrics(ak.pop.nm)
```

## Remove missing loci

A DArT dataset will not have individuals for which the calls are scored all as missing (NA) across all loci, but such individuals may sneak in to the dataset when loci are deleted. Retaining individual or loci with all NAs can cause issues for several functions

```{r}
# Filter for missing loci
ak.pop.nm <- gl.filter.allna(ak.pop.nm)
```

```{r}
nInd(ak.pop.nm)
nLoc(ak.pop.nm)
```

```{r}
gl.smearplot(ak.pop.nm)
```

```{r}
save(ak.pop.nm, file="cache/ak.pop.nm.rdata")
```

## Keep only Magnetic Island subpopulations

```{r}
indNames(ak.pop.nm)
```

```{r}
# Keep only subpopulations from Magnetic Island for fine structure analysis
ak.pop.nm.mag <- gl.keep.pop(ak.pop.nm, pop.list = list("GB", "HB", "HFB", "MB", "MR", "PB", "WB"))
```

```{r}
save(ak.pop.nm.mag, file="cache/ak.pop.nm.mag.rdata")
```

```{r}
nInd(ak.pop.nm.mag)
nLoc(ak.pop.nm.mag)
```

## Convert to structure

Convert ak.pop.nm.mag to structure format to run externally with structure program

```{r}
gl2structure(ak.pop.nm.mag, ind.names = NULL, add.columns = NULL, ploidy = 2, export.marker.names = TRUE, outfile = "ak.pop.mag.nm.str", outpath = "structure", verbose = NULL)
```

```{r}
nInd(ak.pop.nm.mag)
nLoc(ak.pop.nm.mag)
```

## Admixture plot for ak.pop.nm.mag

Calculating BIC for each model as before we see that K=1 is the most preferred model. Nevertheless, we check K=2 just in case this shows any indication of structure related to bays or north/south distinction on Magnetic Island.

```{r}
llpaths_mag <- list.files("structure","k_vs_lik_mag.txt",recursive = TRUE,full.names = TRUE)
reps_mag <- llpaths_mag %>% str_extract("([0-9]+)")

# read_loglik <- function(path,n){
#   read_table(path,col_names = c("k","logLik"))  %>%  
#     mutate(Ldash = logLik - lag(logLik,1)) %>% 
#     mutate(deltaL = lead(Ldash) - Ldash) %>% 
#     add_column(rep=n)
# }


lldata_mag <- map2(llpaths_mag,reps_mag,read_loglik) %>% 
  list_rbind()

lldata_mag %>% 
  group_by(k) %>% 
  summarise(deltaK = mean(abs(deltaL))/var(logLik)) %>% 
  ggplot(aes(x=k,y=deltaK)) + 
    geom_point() + geom_line() + xlab("Number of clusters, K") + ylab("L''")
```

### Load csv

```{r, warning=FALSE}
samples10 <- read_table("structure/ak.pop.mag.nm.samples.txt",col_names = c("ID"))

run10 <- read_table("structure/ak.pop.mag.nm.k3.out.anc.txt",col_names = c("X1", "Label","(%Miss)","C1","C2","C3"),skip = 2) %>% 
  cbind(samples10) %>% 
  select(ID,starts_with("C")) %>% # Select only useful columns
  pivot_longer(-ID,names_to = "cluster",values_to = "p") %>% 
  left_join(metadata,by="ID")
```

For a first glimpse, create a barplot that shows the bays on the x-axis, admixture proportion on y-axis and all clusters stacked.

```{r}

run10 %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x = ID, y = p)) +
  geom_col(aes(fill=cluster),position="stack", stat = "identity",width = 1,linewidth=0) +
  facet_grid(cols=vars(reorder(pop,pop_order)),scales = "free", space = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "bottom") + 
  theme(strip.text = element_text(size=6)) +
  xlab("Sample ID") + ylab("Admixture Proportion")
ggsave("figures/admixture_all_mag.png",width = 3200,height = 800,units = "px")

```

Plots for individual bays are saved into the figures folder

```{r}
# Use lmap to iterate over all pops and run the plotting function on each
admix_plots_by_pop10 <- run10 %>% 
  split(run10$pop) %>% 
  lmap(admix_plot_pop,"10",cluster_colors = c(cluster_colors,"cyan","purple"))
```

Lets use a PCA now to try and understand our result of 3 clusters and see if that matches what we see using a different visualisation technique.

```{r}
load("cache/ak.pop.nm.mag.rdata")

dominant_clusters_run10 <- run10 %>% 
  group_by(ID) %>% 
  slice_max(p) %>% 
  select(ID,cluster,p)

pcoa_ak_filtered_nr_mag <- gl.pcoa(ak.pop.nm.mag, plot.out = FALSE)

pcoa_var_mag <- (pcoa_ak_filtered_nr_mag$eig/sum(pcoa_ak_filtered_nr_mag$eig))[1:2]

pcoa_ak_filtered_nr_meta_mag <-  pcoa_ak_filtered_nr_mag$scores %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  left_join(metadata,by="ID") %>% 
  left_join(dominant_clusters_run10,by="ID")
```

When samples are annotated with their dominant cluster we can see that the three apparent groupings in PCA do indeed correspond with groupings identified by structure. 

```{r}
pcoa_ak_filtered_nr_meta_mag %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=cluster)) + 
  xlab(paste("PC1 (",round(pcoa_var_mag[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var_mag[2]*100,1)," %)",sep=""))
```

Now plotting by location we can see that the largest cluster (C1) is comprised of samples from all bays, however, the two smaller clusters (C3, and C2) are primarily made up of individuals from Picnic Bay and Geoffrey Bay.  Beyond this there is no apparent geographic pattern to genetic structure within the Magnetic Island population. 

```{r}
pcoa_ak_filtered_nr_meta_mag %>% 
  mutate(pop_order = location_order[pop]) %>% 
  ggplot(aes(x=PC1,y=PC2)) + geom_point(aes(color=reorder(pop,pop_order))) + 
  xlab(paste("PC1 (",round(pcoa_var_mag[1]*100,1)," %)",sep="")) +
  ylab(paste("PC2 (",round(pcoa_var_mag[2]*100,1)," %)",sep=""))
```


