---
title: "4 Genetic Diversity"
author: "Sandra Erdmann"
format: gfm
bibliography: references.bib
editor: source
editor_options: 
  chunk_output_type: inline
execute:
  freeze: auto

---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

```{r, echo=FALSE}
## Load Libraries
library(dartR.base)
library(hierfstat)        # allelic richness
library(tidyverse)
source("constants.R")
```

# Genetic Diversity

As a starting point for genetic diversity analyses we use the same filtered set of loci as was used for final population structure analyses. Specifically;

-   These have already been filtered to ensure that they are as reliable as possible while being mindful to avoid filters that would distort allele frequencies and hence bias genetic diversity estimates.  This includes the HWE filter

-   Individuals with admixture proportions that would indicate that they are migrants between MI and "other" populations have been removed. This is because these these individuals have uncertain origins (ie parents likely from a different reef to where they were found).

-   Highly admixed individuals (admixture proportion \> 5%) between the two major lineages in this study (ie Maggie vs surrounding reefs) have been excluded. See note in the section on population structure about this threshold and the relatively small number of individuals it excludes.

-   Individuals that are likely genetic clones or close relatives have been removed as these are undesirable in pretty much all analyses.

The starting data satisfying all these criteria is called `ak.pop.nm` as was created in the `03.population_structure` script

```{r}
# Loads ak.pop.nm
load("cache/ak.pop.nm.rdata")
```

## Population groupings

To facilitate diversity statistics we use population groupings following both known genetic structure and geography. Datasets reflecting these groupings are;

| Dataset Name    | Grouping                                                  |
|-------------------------------------|-----------------------------------|
| `ak.pop.nm`     | All sites grouped separately                              |
| `ak.gen.no.so`  | Magnetic Island grouped into North and South              |
| `ak.gen.mi.adj` | All sites grouped into Magnetic Island and Adjacent Reefs |

```{r, results='hide'}
# Creation of the `ak.gen.no.so` dataset

if (!file.exists("cache/ak.gen.no.so.rdata")) {
  # Merge subpopulations from north Magnetic Island into one population
  ak.gen.no.so <- gl.merge.pop(ak.pop.nm, old=c("HB", "HFB", "MB", "WB"), new='no')
  # Using this merged dataset, merge also subpopulations from south Magnetic Island into one population. 
  ak.gen.no.so <- gl.merge.pop(ak.gen.no.so, old=c("GB", "MR", "PB"), new='so')
  ak.gen.no.so <- gl.merge.pop(ak.gen.no.so, old=c("SO", "LPB", "EP", "WP", "HaR","BRR", "JB", "KR"), new='ADJ')
  ak.gen.no.so <- gl.filter.monomorphs(ak.gen.no.so)
  ak.gen.no.so <- gl.recalc.metrics(ak.gen.no.so)
  save(ak.gen.no.so,file = "cache/ak.gen.no.so.rdata")
} else {
  load("cache/ak.gen.no.so.rdata")
}
```

```{r, results='hide'}
if (!file.exists("cache/ak.gen.mi.adj.rdata")){
  ak.gen.mi.adj <- gl.merge.pop(ak.pop.nm, old=c("GB", "HB", "HFB", "MB", "MR", "PB", "WB"), new='MI')
  ak.gen.mi.adj <- gl.merge.pop(ak.gen.mi.adj, old=c("SO", "LPB", "EP", "WP", "HaR","BRR", "JB", "KR"), new='ADJ')
  ak.gen.mi.adj <- gl.filter.monomorphs(ak.gen.mi.adj)
  ak.gen.mi.adj <- gl.recalc.metrics(ak.gen.mi.adj)
  save(ak.gen.mi.adj,file = "cache/ak.gen.mi.adj.rdata")
  write_rds(ak.gen.mi.adj,file = "basic_stats/ak.gen.mi.adj.rds")
} else {
  load("cache/ak.gen.mi.adj.rdata")
}

```

## Convert to hierfstat format

```{r,results='hide'}
# Convert genlight objects to genind format
ak.pop.nm.hfstat <- gl2gi(ak.pop.nm) %>% genind2hierfstat()
ak.gen.no.so.hfstat <- gl2gi(ak.gen.no.so)  %>% genind2hierfstat() 
ak.gen.mi.adj.hfstat <- gl2gi(ak.gen.mi.adj) %>% genind2hierfstat()
write_rds(list(ak.pop.nm.hfstat,ak.gen.no.so.hfstat,ak.gen.mi.adj.hfstat),"basic_stats/hfstat.rds")
```

# Measures of Genetic Diversity as EBVs

Genetic Diversity is one of four EBVs (Essential Biodiversity Variables) proposed by [@Hoban2022-ek]. It is composed of richness and evenness. Here, we calculate Allelic Richness (Ar) as a measure of richness and expected heterozygosity (He) as a measure of evenness.

## Allelic richness

Our data comprise a set of biallelic SNPs which means that the maximum number of alleles at any locus is 2. Allelic richness for the entire dataset with no subdivisions should be exactly 2 since every SNP must have exactly 2 alleles in order to be included. Values of AR less than 2 come from the fact that some alleles will be missing when a subsample is made from the whole. The larger the sample the higher the value of AR so the `allelic.richness` function from hierfstat is used here. This calculates a rarefied value (accounting for differences in sample size between populations).

```{r}
# Calculate allelic richness
ar <- allelic.richness(ak.pop.nm.hfstat)
ar.no.so <- allelic.richness(ak.gen.no.so.hfstat)
ar.mi.adj <- allelic.richness(ak.gen.mi.adj.hfstat)
```

```{r}
load("cache/sites_data.rdata")
```

```{r}
summarise_pops <- function(coltable){
  coltable %>% 
  colMeans(na.rm = TRUE) %>% 
  as.data.frame() %>% 
  rownames_to_column("pop") %>% 
  left_join(sites_data)
}
```


Since Ar is heavily impacted by sample size we avoid super small samples on individual reefs and report population groupings such as "Adjacent Reefs",  North and South Maggie.

```{r}
ar_means <- (ar.no.so$Ar %>% 
  colMeans(na.rm = TRUE))
data.frame(pop=names(ar_means),Ar = ar_means) %>% knitr::kable()
```

To formally test whether Ar is higher at adjacent than at Magnetic Island we use a permutation test. Here this is implemented by randomly reassigning individuals to populations.  To perform this shuffling we ran 1000 iterations with an R script [basic_stats/run_permute.R](run_permute.R) to perform each iteration.

```{r}
if ( file.exists("cache/perm_data.rds")){
# This reads all permutation data. Including for basic stats below
  perm_data <- list.files("basic_stats/perms/",pattern = ".rds",full.names = TRUE) %>% 
    map(read_rds)
  write_rds(perm_data,"cache/perm_data.rds")
} else {
  perm_data <- read_rds("cache/perm_data.rds")
}
```

```{r}
m_vs_adj_actual_value <- ar.mi.adj$Ar %>% colMeans(na.rm = TRUE) %>% diff()
mi_adj_ar <- perm_data %>% map(~ colMeans(.x$adj.mi$Ar$Ar)) %>% 
  bind_rows() %>% 
  mutate(diff = abs(MI-ADJ)) 

p_mi_adj_ar <- (mi_adj_ar %>% 
  filter(abs(diff)>abs(m_vs_adj_actual_value)) %>% nrow())/1000

```

```{r}
no_vs_so_actual_value <- (ar.no.so$Ar %>% colMeans(na.rm = TRUE))[c("no","so")] %>% diff()
no_so_adj_ar <- perm_data %>% map(~ colMeans(.x$no.so$Ar$Ar)) %>% 
  bind_rows() %>% 
  mutate(no_vs_so = abs(no-so)) %>% 
  mutate(diff = abs(no-so))

p_no_so_ar <- (no_so_adj_ar %>% 
  filter(diff>no_vs_so_actual_value) %>% nrow())/1000
```

This test reveals that the difference between MI and adjacent is highly significant (p<0.001) but for the contrast between north and south magnetic Island it is not.

```{r}
data.frame(contrast = c("MI vs ADJ","NO vs SO"), "Ar Diff" = c(m_vs_adj_actual_value,no_vs_so_actual_value), "p-value"=c(p_mi_adj_ar,p_no_so_ar)) %>% knitr::kable()
```


## Heterozygosity

Next we used the `basic.stats` function in `hierfstat` to calculate expected heterozygosity (He, also called Hs), observed heterozygosity (Ho) and Fis.  `He/Hs` is a measure of genetic evenness.  `Fis` is a measure of heterozygote deficiency.  We see that all populations show a background level of `Fis`

```{r}
basic_stats <- basic.stats(ak.pop.nm.hfstat)
basic_stats_adj.no.so <- basic.stats(ak.gen.no.so.hfstat)
basic_stats_adj.mi <- basic.stats(ak.gen.mi.adj.hfstat)
```


```{r}
# Assemble all stats into a table averaged by column
assemble_basic_stats <- function(bs){
  ho.stats <- bs$Ho %>% 
    summarise_pops() %>% 
    dplyr::rename(Ho = ".") 

  he.stats <- bs$Hs %>% 
    summarise_pops() %>% 
    dplyr::rename(He = ".") 

  fis.stats <- bs$Fis %>% 
    summarise_pops() %>% 
    dplyr::rename(fis = ".") 

  ho.stats %>% 
    left_join(he.stats) %>% 
    left_join(fis.stats) %>% 
    pivot_longer(all_of(c("He","Ho","fis")), names_to = "stat", values_to = "stat_value")
}

# Gives aggregated values for three main population groupings
allhet.stats_noso <- assemble_basic_stats(basic_stats_adj.no.so) %>% 
  select(no_so_group = pop,stat,stat_value_agg = stat_value)

# Collects aggregated and subdivided values together
allhet.stats <- assemble_basic_stats(basic_stats) %>% 
  mutate(no_so_group = case_when(
    pop_group == "ADJ" ~ "ADJ",
    pop %in% c("PB","GB","MR") ~ "so",
    .default = "no"
  )) %>% left_join(allhet.stats_noso)
  
```

First plot all stats subdivided by reef

```{r}
allhet.stats %>% 
  ggplot(aes(x=reorder(pop,pop_order),y=stat_value)) + 
  geom_point(aes(color=pop_group)) + 
  facet_wrap(~stat, scales="free", ncol=1) +
  geom_point(aes(color=pop_group)) + 
  scale_color_manual(values = pop_colors, labels = pop_names) +
  labs(
    x="Reef",
    y="Stat Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top", legend.title = element_blank()) 
```

Alternatively we can present this as a boxplot

```{r}
allhet.stats %>% 
  ggplot(aes(x=no_so_group,y=stat_value)) + 
  geom_boxplot(aes(color=no_so_group)) + 
  geom_point(aes(y=stat_value_agg)) +
  facet_wrap(~stat, scales = "free") +
#  scale_color_manual(values = pop_colors, labels = pop_names) +
  labs(
    x="",
    y="Stat Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top", legend.title = element_blank()) 
```

Permutation tests were also used to calculate significance of differences for these statistics.  This is done on aggregated data, grouped into ADJ, NO and SO.  Note the somewhat unexpected result that the smaller populations at Magnetic Island show slightly higher heterozygosity and lower Fis than the larger Adjacent population.  Note that this effect is numerically very small, and likely reflects expected heterozygote excess in small populations due to differences in allele frequencies between sexes as a result of binomial sampling error.  This effect applies equally to self-incompatible hermaphroditic species [@Balloux2004-yn] such as Acroporid corals and forms the basis of the heterozygote excess method for calculating Ne [@Luikart1999-yh]

```{r}
test_basic_stat <- function(stat){
  actual_values <- basic_stats_adj.mi[[stat]] %>% colMeans(na.rm = TRUE)
  
  mi_adj <- perm_data %>% map(~ colMeans(.x[["adj.mi"]][["basic_stats"]][[stat]])) %>% 
    bind_rows() %>% 
    mutate(diff = abs(MI-ADJ)) 

  p_mi_adj <- (mi_adj %>% 
    filter(abs(diff)>abs(diff(actual_values))) %>% nrow())/1000  
  data.frame(stat,values=paste(actual_values, collapse = ";"),diff = diff(actual_values), "p"=p_mi_adj)
}

test_basic_stat_noso <- function(stat){
  actual_values <- basic_stats_adj.no.so[[stat]] %>% colMeans(na.rm = TRUE)
  
  data_withdiffs <- perm_data %>% map(~ colMeans(.x[["no.so"]][["basic_stats"]][[stat]],na.rm = TRUE)) %>% 
    bind_rows() %>% 
    mutate(adj_no = abs(ADJ-no), adj_so = abs(ADJ-so),no_so = abs(no-so)) 

  pval_adj_no <- (data_withdiffs %>% 
    filter(abs(adj_no)>abs(diff(actual_values[c("ADJ","no")]))) %>% nrow())/1000  

  pval_adj_so <- (data_withdiffs %>% 
    filter(abs(adj_so)>abs(diff(actual_values[c("ADJ","so")]))) %>% nrow())/1000

  pval_no_so <- (data_withdiffs %>% 
    filter(abs(no_so)>abs(diff(actual_values[c("no","so")]))) %>% nrow())/1000
  
  data.frame(stat,contrast = c("adj_no","adj_so","no_so"), "p"=c(pval_adj_no,pval_adj_so,pval_no_so))
}

p_values <- map(c("Hs","Ho","Fis"),test_basic_stat_noso) %>% 
  bind_rows() %>% 
  pivot_wider(names_from = stat,values_from = p)
knitr::kable(p_values)

allhet.stats_noso %>% knitr::kable()
```


## Fixed allelic differences/Private alleles

If we look at private alleles and fixed differences between all pairs of reefs we see that the number of private alleles strongly reflects the sample size. We also see a few examples of fixed differences (alternate alleles fixed in different populations) particularly betwen WP and Magnetic Island reefs.

```{r}
pa_ak.gen <-dartR::gl.report.pa(ak.pop.nm)
```

When running this on amalgamated populations we now see no fixed differences between MI and ADJ but plenty of private alleles. Interestingly if we break up MI into no and so there are 2 loci fixed between no and ADJ.

```{r}
pa_ak.gen.mi <- gl.report.pa(ak.gen.mi.adj)
pa_ak.gen.no.so <- gl.report.pa(ak.gen.no.so)
```

# 
