---
title: "5 Ne Estimates"
author: "Sandra Erdmann and Ira Cooke"
format: gfm
editor: source
editor_options: 
  chunk_output_type: inline
execute:
  freeze: auto
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

```{r, echo=FALSE}
## Load Libraries
library(tidyverse)
library(dartRverse)
#library(dartR.base)
#library(dartR.popgen)
library(vcfR)
source("constants.R")
```

# Effective Population Size

Here we use the LD method implemented in Neestimator2 to estimate effective population size. This method is based on the fact that as Ne becomes small even a perfect random-mating population will experience some inbreeding and this will induce LD. A key assumption of the method is that the pairs of loci used for this calculation are on separate chromosomes because we do not want to include LD due to physical linkage.

Although the original Neestimator2 program doesn't appear to be maintained (DNS error attempting to access <http://www.molecularfisherieslaboratory.com.au/>) a binary is distributed with the [RLDNe software](https://github.com/zakrobinson/RLDNe)

Sources of bias

- Age structure downward bias as it can introduce a form of Wahlund effect [@Waples2014-ge]
- Similarly, individuals from distinct populations are combined into the same population this is a classic Wahlund effect and will also downwardly bias Ne estimates [@Kardos2025-ol]
- Migration can cause upward bias (due to additional parents) and downward bias (due to mixture LD), however a simulation study by [@Waples2011-vd] showed that the LDNe method is generally robust to migration as long as it is less than about 5-10\% per generation.

Thus we start by installing this package

```{r, eval=FALSE}
devtools::install_github(repo="zakrobinson/RLDNe")
```

The path to the NE estimator binary can be obtained as follows (adjust if not using a mac)

```{r}
neest_path = paste(find.package("RLDNe"),"bin/Mac",sep = "/")
```

Choice of loci and individuals for Ne estimation follows similar principles to that of genetic diversity estimates. Thus we can use the `ak.pop.nm` dataset as a starting point.

```{r}
# Loads ak.pop.nm
load("cache/ak.pop.nm.rdata")
load("cache/metadata.rdata")
```

## Adding chromosome information

To avoid pairing loci on the same chromosome we need to include chromosome info in our genlight object. This info is available in the vcf generated via `dart2vcf` in [01.data_wrangling_and_filtering](01.data_wrangling_and_filtering.md). To include it we need to generate a vcf with only our loci and individuals from the `ak.pop.nm` data.

```{r, eval=FALSE}
write.table(ak.pop.nm$loc.names,file = "dart2vcf/ak.pop.nm.loci",row.names = FALSE,quote = FALSE)
write.table(ak.pop.nm$ind.names,file = "dart2vcf/ak.pop.nm.ind",row.names = FALSE,quote = FALSE,col.names = FALSE)
```

A vcf with these loci and individuals is then created using the script `dart2vcf/06.filter3.sh`

```{r, results='hide'}
ak.chr <- dartR::gl.read.vcf("dart2vcf/ak.pop.nm.ind.vcf.gz", ind.metafile = "ind_metrics_SNP2_namefix.csv")
```

## Merging populations

Estimation of Ne from LD information benefits from larger sample sizes, especially when the actual value of Ne is large. We therefore merged populations into the following units

| Merged Unit | Sampling Locations             | Sample Size |
|-------------|--------------------------------|-------------|
| no          | "HB", "HFB", "MB", "WB"        | 59          |
| so          | "GB", "MR", "PB"               | 58          |
| MS          | "BRR", "JB", "KR"              | 48          |
| PI          | "SO", "LPB", "EP", "HaR", "WP" | 238         |

Note WP is removed from the PI population because these individuals were slightly divergent in Fst analyses.

```{r, results='hide'}
#wp.inds <- metadata %>% filter(pop=="WP") %>% pull(ID)

# Merge subpopulations from north Magnetic Island into one population
ak.chr.no.so <- gl.merge.pop(ak.chr, old=c("HB", "HFB", "MB", "WB"), new='no')

# Using this merged dataset, merge also subpopulations from south Magnetic Island into one population. 
ak.chr.no.so <- gl.merge.pop(ak.chr.no.so, old=c("GB", "MR", "PB"), new='so')

# Merge Mid-shelf
ak.chr.no.so <- gl.merge.pop(ak.chr.no.so, old=c("BRR", "JB", "KR"), new='MS')

# Merge PI
ak.chr.no.so <- gl.merge.pop(ak.chr.no.so, old=c("SO", "LPB", "EP", "HaR", "WP"), new='PI')

ak.chr.no.so <- gl.filter.monomorphs(ak.chr.no.so)
ak.chr.no.so <- gl.recalc.metrics(ak.chr.no.so)
```

```{r, results='hide'}
# And then make another dataset where adjacent sites are all merged
ak.chr.mi.adj <- gl.merge.pop(ak.chr.no.so, old=c("PI","MS"), new='adj')
ak.chr.mi.adj <- gl.filter.monomorphs(ak.chr.mi.adj)
ak.chr.mi.adj <- gl.recalc.metrics(ak.chr.mi.adj)
write_rds(ak.chr.mi.adj,"ne/ak.chr.mi.adj.rds")
```

### Ne all merged units

```{r}
if ( !file.exists("cache/ne.chr.no.so.rdata") ){
  ne.chr.no.so <- dartR.popgen::gl.LDNe(ak.chr.no.so, outfile = "LD.chr.no.so.nm.txt", neest.path = neest_path,critical = c(0, 0.05),plot.out = FALSE,pairing = "separate")

  save(ne.chr.no.so,file = "cache/ne.chr.no.so.rdata")
} else {
  load("cache/ne.chr.no.so.rdata")
}
```

Ne estimates for Midshelf (almost offshore) reefs are astronomically high and upper confidence intervals are Inf (not estimable). This is probably due to a combination of low sample size and high Ne leading to poor estimates. Therefore we combine MS and PI into one population (Adjacent reefs).

### Ne North and South Magnetic Island vs Adjacent reefs

```{r}
if (!file.exists("cache/ne.chr.mi.adj.rdata")){

  ne.chr.mi.adj <- dartR.popgen::gl.LDNe(ak.chr.mi.adj,outfile = "LD.chr.mi.adj.nm.txt", neest.path =  neest_path,critical = c(0, 0.05),plot.out = FALSE,pairing = "separate")

  save(ne.chr.mi.adj, file = "cache/ne.chr.mi.adj.rdata")
} else {
  load("cache/ne.chr.mi.adj.rdata")
}
```

### Ne Magnetic Island (merged) vs Adjacent reefs

```{r, results='hide'}
# And then make another dataset where adjacent sites are all merged
ak.chr.mi.merged.adj <- gl.merge.pop(ak.chr.mi.adj, old=c("no","so"), new='MI')
ak.chr.mi.merged.adj <- gl.filter.monomorphs(ak.chr.mi.merged.adj)
ak.chr.mi.merged.adj <- gl.recalc.metrics(ak.chr.mi.merged.adj)
```

```{r}
if (!file.exists("cache/ne.chr.mi.merged.adj.rdata")){

  ne.chr.mi.merged.adj <- dartR.popgen::gl.LDNe(ak.chr.mi.merged.adj, outfile = "LD.chr.mi.merged.adj.nm.txt", neest.path =  neest_path,critical = c(0, 0.05),plot.out = FALSE,pairing = "separate")

  save(ne.chr.mi.merged.adj, file = "cache/ne.chr.mi.merged.adj.rdata")
} else {
  load("cache/ne.chr.mi.merged.adj.rdata")
}
```

```{r}
# Helper function for extracting Ne results.
extract_ne_data <- function(netable){
  f1 <- as.numeric(netable[,2]) # >0.05 
  # Corresponds to Ne, Lower CI, Upper CI
  data.frame(ne=f1[6],ne_min=f1[7],ne_max=f1[8])
}

ne_results <- map(ne.chr.mi.adj,extract_ne_data) %>% 
  list_rbind(names_to = "pop") %>%  
  mutate(pop = case_when(
    pop=="adj" ~ "ADJ",
    grepl("^no",pop) ~ "NO",
    grepl("^so",pop) ~ "SO"
  )) %>% 
  mutate(pop_name = pop_names[pop])
```

```{r}
ggplot(ne_results,aes(x=pop_name,y=ne,ymin=ne_min,ymax=ne_max)) + 
  geom_crossbar(aes(color = pop), linewidth = 1) +
  scale_color_manual(values = pop_colors) +
  labs(
    x = "",
    y= "Ne"
  ) +
  scale_y_log10() + 
  theme_minimal() + 
  theme(legend.position = "None")
```

```{r}
ne_results %>% knitr::kable()
```

The NEEstimator software provides parametric and pseudo-jackknife methods to provide confidence intervals on Ne estimates. Since we have plenty of computational power we performed a full Jackknife on samples (removing each sample successively). This gives wider confidence intervals but it probably more realistic.  The Jackknife estimate required running LDNe 403 times (one for each sample).  Scripts to submit each sample as a standalone job are found in the [ne](ne) subdirectory.

```{r}
read_ne <- function(path){
  bn=basename(path) %>% str_remove(".rds")
  read_rds(path)[[1]][,c(1,2)] %>% 
    add_column(input=bn)
}

sample_sizes <- as.vector(table(ak.chr.mi.adj$pop))
names(sample_sizes) <- c("adj","so","no")
```

```{r}
# This provides parameters for each job when running the jackknife
data.frame(id=ak.chr.mi.adj$ind.names,pop=ak.chr.mi.adj$pop) %>% 
  mutate(iteration = row_number()) %>% 
  write_tsv(file = "ne/grid_jk.tsv",col_names = FALSE)
```

```{r}
ne_grid_jk <- list.files("ne/jackknife",pattern = "[0-9].rds",full.names = TRUE) %>% 
  map(read_ne) %>% 
  bind_rows() %>% 
  separate(input,into = c("pop","rep"),sep="_") %>% 
  filter(Statistic=="Estimated Ne^") %>% 
  mutate(No_S=as.numeric(`Frequency 1`),rep=as.integer(rep))

ne_jk_summary <- ne_grid_jk %>% 
  mutate(lgne = log10(No_S)) %>% # We log transform before applying the Jackknife because this produces better variance estimates
  group_by(pop) %>% 
  summarise(
    jkmean = mean(lgne),
    jkvar = (n()-1)*var(lgne),
    jk_se=sqrt(jkvar),
    jk_low=jkmean-1.96*jk_se,
    jk_high = jkmean+1.96*jk_se
    ) 
  
  
```

```{r}
ne_grid_jk %>% 
  ggplot(aes(x=pop,y=No_S)) + 
  geom_violin() +
  geom_point() + 
  scale_y_log10()
```

```{r}
ne_jk_summary %>% 
ggplot(aes(x=pop,y=10^jkmean,ymin=10^jk_low,ymax=10^jk_high)) + 
  geom_crossbar(aes(color = pop), linewidth = 1) + 
  scale_y_log10()
```
