---
title: "10 Sensitivity of genetic diversity to close kin"
author: "Ira Cooke"
format: gfm
editor: source
editor_options:
  chunk_output_type: inline
execute:
  freeze: auto
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = FALSE)
```

```{r libraries}
library(dartR.base)
library(hierfstat)
library(tidyverse)
source("constants.R")
options(readr.show_col_types = FALSE)
```

The genetic diversity analyses in `04.genetic_diversity.qmd` are based on
`ak.pop.nm`, a dataset from which duplicates had been removed but not close kin. Here we recompute allelic richness (Ar), observed and expected heterozygosity
(Ho, Hs) and Fis with kin removed. 

```{r kin-list}
excluded_kin <- read_tsv("relatedness/excluded_kin.tsv") |>
  pull(excluded_kin) |>
  unique()

excluded_kin
```

## Datasets

We use `mono.rm = FALSE` so that both datasets retain
exactly the same loci. If monomorphic loci were dropped after kin removal the
two analyses would be averaging over different locus sets, which would itself
change mean diversity irrespective of which individuals were included.  Either way the effect is likely tiny since we are talking about just 5 SNPs that are monomorphic after removing kin.

```{r datasets, results='hide'}
load("cache/ak.pop.nm.rdata")

ak.div.all <- ak.pop.nm

ak.div.nokin <- gl.drop.ind(
  ak.div.all,
  ind.list = intersect(excluded_kin, indNames(ak.div.all)),
  mono.rm = FALSE, recalc = TRUE
)
```

```{r loci-check}
c(all = nLoc(ak.div.all), nokin = nLoc(ak.div.nokin))
```

Seven of the additionally removed individuals come from two southern Magnetic Island
bays, Geoffrey Bay (GB) and Picnic Bay (PB), and just one comes from the adjacent population.

```{r sample-sizes}
knitr::kable(
  full_join(
    as.data.frame(table(pop(ak.div.all))),
    as.data.frame(table(pop(ak.div.nokin))),
    by = "Var1", suffix = c("_all", "_nokin")
  ) |>
    rename(site = Var1, n_all = Freq_all, n_nokin = Freq_nokin) |>
    mutate(n_removed = n_all - n_nokin)
)
```

## Population groupings

We report statistics for the three main groupings (adjacent reefs, north and south Magnetic Island).
Merging is done without filtering monomorphs, again to keep the locus set fixed.

```{r groupings, results='hide'}
adjacent_sites <- setdiff(location_codes, maggie_sites)

merge_no_so <- function(gl) {
  gl |>
    gl.merge.pop(old = maggie_no_sites, new = "no") |>
    gl.merge.pop(old = maggie_so_sites, new = "so") |>
    gl.merge.pop(old = adjacent_sites, new = "ADJ")
}

merge_adj_mi <- function(gl) {
  gl |>
    gl.merge.pop(old = maggie_sites, new = "MI") |>
    gl.merge.pop(old = adjacent_sites, new = "ADJ")
}

# The all-individuals hierfstat objects were written by 04.genetic_diversity.qmd.
# Reading them rather than rebuilding guarantees that the observed statistics
# here are computed on exactly the objects the permutation jobs were given.
hfstat_all <- read_rds("basic_stats/hfstat.rds")

# The equivalent kin-removed objects, in the same list order, written out to
# serve as input to the permutation jobs.
if (file.exists("basic_stats/hfstat_nokin.rds")) {
  hfstat_nokin <- read_rds("basic_stats/hfstat_nokin.rds")
} else {
  hfstat_nokin <- list(
    ak.div.nokin,
    merge_no_so(ak.div.nokin),
    merge_adj_mi(ak.div.nokin)
  ) |>
    map(function(gl) gl2gi(gl) |> genind2hierfstat())

  write_rds(hfstat_nokin, "basic_stats/hfstat_nokin.rds")
}

names(hfstat_all) <- c("site", "no.so", "adj.mi")
names(hfstat_nokin) <- c("site", "no.so", "adj.mi")
```

Both sets of objects carry the same loci, and differ only by the eight
individuals.

```{r grouping-check}
knitr::kable(
  tibble(
    grouping = names(hfstat_all),
    n_ind_all = map_int(hfstat_all, nrow),
    n_ind_nokin = map_int(hfstat_nokin, nrow),
    n_loci = map_int(hfstat_nokin, function(x) ncol(x) - 1L)
  )
)
```

## Diversity statistics

`allelic.richness` rarefies to `2 * min(ind.count(data))`, i.e. the smallest
number of genotyped individuals seen in any population at any locus. Since we would like to compare results with/without kin we need to make sure we set this to the same value in each analysis.  We therefore fix `min.n` to a common value across the two datasets being compared.  This means re-running the analysis including kin (with a slightly lower min.n) and running analysis without kin at the same min.n

```{r stats}
pop_means <- function(hf, min_n) {
  bs <- basic.stats(hf)
  list(
    Ar = allelic.richness(hf, min.n = min_n)$Ar,
    Ho = bs$Ho,
    Hs = bs$Hs,
    Fis = bs$Fis
  ) |>
    # Each element is a loci x populations matrix; average over loci
    map(function(m) colMeans(m, na.rm = TRUE) |> enframe(name = "pop")) |>
    bind_rows(.id = "stat")
}

compare_diversity <- function(hf_all, hf_nokin) {
  min_n <- min(
    2 * min(ind.count(hf_all), na.rm = TRUE),
    2 * min(ind.count(hf_nokin), na.rm = TRUE)
  )

  bind_rows(
    "All individuals" = pop_means(hf_all, min_n),
    "Kin removed" = pop_means(hf_nokin, min_n),
    .id = "dataset"
  ) |>
    mutate(min_n = min_n)
}

div_site <- compare_diversity(hfstat_all$site, hfstat_nokin$site)
div_group <- compare_diversity(hfstat_all$no.so, hfstat_nokin$no.so)
div_adj_mi <- compare_diversity(hfstat_all$adj.mi, hfstat_nokin$adj.mi)
```

```{r rarefaction-depth}
c(
  site = unique(div_site$min_n),
  no.so = unique(div_group$min_n),
  adj.mi = unique(div_adj_mi$min_n)
)
```

### Three main groupings

```{r group-table}
compare_table <- function(div) {
  div |>
    select(dataset, stat, pop, value) |>
    pivot_wider(names_from = dataset, values_from = value) |>
    mutate(difference = `Kin removed` - `All individuals`) |>
    arrange(stat, pop)
}

knitr::kable(compare_table(div_group), digits = 4)
```

```{r}
compare_table(div_group) |>
  mutate(stat = ifelse(stat=="Hs","He",stat)) |> 
  pivot_longer(c(3,4)) |> 
  ggplot(aes(x=pop,y=value)) + 
  geom_point(aes(color=name),size=3,alpha=0.8) + 
  facet_wrap(~stat,scales="free") + 
  theme_bw() + 
  ylab("") + xlab("") + theme(legend.position = "bottom",legend.title = element_blank())
ggsave("figures/figure_s15.png",width = 8,height = 5)
```


Note that these Ar values differ slightly from those reported in
`04.genetic_diversity.qmd` because they are computed at a common rarefaction depth
(`r unique(div_group$min_n)` alleles) across kin-removed and kin-included rather than at the slightly higher depth used in the main analysis.

## Permutation tests

Significance of the differences between groupings was assessed by permutation,
randomly reassigning individuals to populations and recomputing each statistic.
Both datasets were run as 1000-iteration slurm array jobs on an external HPC
system using the scripts in `basic_stats`:

| Slurm script | Worker | Input | Output |
|:--|:--|:--|:--|
| `01.ar_perm.sh` | `run_permute.R` | `hfstat.rds` | `perms/` |
| `02.ar_perm_nokin.sh` | `run_permute_nokin.R` | `hfstat_nokin.rds` | `perms_nokin/` |
| `03.ar_perm_all.sh` | `run_permute_all.R` | `hfstat.rds` | `perms_all/` |

We use `perms_all` and `perms_nokin` here. These differ from the original `perms/`
in one respect: the allelic richness rarefaction depth is fixed at a value common
to both datasets rather than the slightly higher value if only the full dataset is being considered. 

```{r read-perms}
#Each permutation file holds full loci by population matrices and the two
#directories total around 7 GB, so we reduce each file to per-population means as
#it is read.

if (file.exists("cache/perm_summary_kin.rds")) {
  perm_summary <- read_rds("cache/perm_summary_kin.rds")
} else {
  summarise_perm_file <- function(path) {
    p <- read_rds(path)
    imap(p, function(g, gname) {
      list(
        Ar = colMeans(g$Ar$Ar, na.rm = TRUE),
        Ho = colMeans(g$basic_stats$Ho, na.rm = TRUE),
        Hs = colMeans(g$basic_stats$Hs, na.rm = TRUE),
        Fis = colMeans(g$basic_stats$Fis, na.rm = TRUE)
      ) |>
        map(function(v) enframe(v, name = "pop")) |>
        bind_rows(.id = "stat")
    }) |>
      bind_rows(.id = "grouping")
  }

  summarise_perm_dir <- function(dir) {
    list.files(dir, pattern = "\\.rds$", full.names = TRUE) |>
      set_names(function(p) tools::file_path_sans_ext(basename(p))) |>
      map(summarise_perm_file) |>
      bind_rows(.id = "rep")
  }

  perm_summary <- bind_rows(
    "All individuals" = summarise_perm_dir("basic_stats/perms_all"),
    "Kin removed" = summarise_perm_dir("basic_stats/perms_nokin"),
    .id = "dataset"
  )

  write_rds(perm_summary, "cache/perm_summary_kin.rds")
}
```


The test statistic is the difference in the locus-averaged statistic between two
populations. The p-value is the proportion of permutations whose absolute
difference is at least as large as the observed one, computed as
`(n_extreme + 1) / (n_perm + 1)` so that the smallest attainable value is 1/1001
rather than zero.

```{r perm-tests}
observed_means <- bind_rows(
  no.so = div_group,
  adj.mi = div_adj_mi,
  .id = "grouping"
) |>
  select(dataset, grouping, stat, pop, value)

group_contrasts <- tribble(
  ~grouping, ~contrast,  ~pop_a, ~pop_b,
  "adj.mi",  "ADJ - MI", "ADJ",  "MI",
  "no.so",   "ADJ - no", "ADJ",  "no",
  "no.so",   "ADJ - so", "ADJ",  "so",
  "no.so",   "no - so",  "no",   "so"
)

contrast_diffs <- function(means) {
  means |>
    inner_join(group_contrasts, by = "grouping", relationship = "many-to-many") |>
    filter(pop == pop_a | pop == pop_b) |>
    mutate(side = if_else(pop == pop_a, "a", "b")) |>
    select(any_of(c("dataset", "rep")), grouping, contrast, stat, side, value) |>
    pivot_wider(names_from = side, values_from = value) |>
    mutate(diff = a - b) |>
    select(-a, -b)
}

perm_tests <- contrast_diffs(perm_summary) |>
  rename(perm_diff = diff) |>
  left_join(
    contrast_diffs(observed_means) |> rename(obs_diff = diff),
    by = c("dataset", "grouping", "contrast", "stat")
  ) |>
  group_by(dataset, grouping, contrast, stat) |>
  summarise(
    obs_diff = first(obs_diff),
    n_perm = n(),
    n_extreme = sum(abs(perm_diff) >= abs(obs_diff)),
    p = (n_extreme + 1) / (n_perm + 1),
    .groups = "drop"
  )

perm_test_table <- function(s) {
  perm_tests |>
    filter(stat == s) |>
    select(grouping, contrast, dataset, obs_diff, p) |>
    arrange(grouping, contrast, dataset) |>
    knitr::kable(digits = 4)
}
```

### Allelic richness

```{r ar-tests}
perm_test_table("Ar")
```

Adjacent reefs carry significantly higher allelic richness than either Magnetic
Island grouping, while north and south Magnetic Island are not distinguishable
from each other. Both conclusions hold whether or not close kin are included, and
the observed differences are essentially unchanged.

```{r ar-null-plot, fig.width=9, fig.height=5}
contrast_diffs(perm_summary) |>
  filter(stat == "Ar") |>
  ggplot(aes(x = diff)) +
  geom_histogram(bins = 40) +
  geom_vline(
    data = contrast_diffs(observed_means) |> filter(stat == "Ar"),
    aes(xintercept = diff),
    linetype = "dashed"
  ) +
  facet_grid(dataset ~ contrast, scales = "free") +
  theme_bw() +
  labs(x = "Difference in mean Ar", y = "Permutations")
```

### Heterozygosity and Fis

Observed Heterozygosity (Ho)
```{r het-tests}
perm_test_table("Ho")
```

Expected Heterozygosity (He)

```{r}
perm_test_table("Hs")
```

Fis

```{r}
perm_test_table("Fis")
```

These results show the same conclusions (based on a p-value threshold of 0.05) as were found in `04.genetic_diversity`.  Every
contrast involving observed heterozygosity (Ho) is significant, including north
versus south Magnetic Island, and in each case Magnetic Island is the *higher*
value. None of the expected heterozygosity (Hs) contrasts reach significance. Fis is significantly lower at Magnetic Island than at adjacent reefs.

## Summary

```{r summary-values, include=FALSE}
max_abs_diff <- compare_table(div_site) |>
  group_by(stat) |>
  summarise(max_abs_diff = max(abs(difference))) |>
  deframe()

changed_sites <- compare_table(div_site) |>
  filter(abs(difference) > 1e-10) |>
  pull(pop) |>
  unique()

# Size of the ADJ vs Magnetic Island contrast, for comparison with the kin effect
ar_gap <- compare_table(div_group) |>
  filter(stat == "Ar") |>
  summarise(gap = max(`Kin removed`) - min(`Kin removed`)) |>
  pull(gap)

# Keep small values out of scientific notation, which renders badly in gfm
fmt <- function(x, digits = 2) format(signif(x, digits), scientific = FALSE)

# Do the two datasets agree on which contrasts are significant at 0.05?
sig_agreement <- perm_tests |>
  select(dataset, grouping, contrast, stat, p) |>
  pivot_wider(names_from = dataset, values_from = p) |>
  mutate(agree = (`All individuals` < 0.05) == (`Kin removed` < 0.05))
```

```{r}
sig_agreement |> knitr::kable()
```

